---
title: "Workshop R spatial"
author: "Team inSileco"
date: "2026-03-25"
from: markdown+emoji
format: revealjs
---

```{=html}
<script>
document.addEventListener('DOMContentLoaded', function() {
  var logo = document.querySelector('.slide-logo');
  if (logo) {
    var link = document.createElement('a');
    link.href = 'https://www.insileco.io';
    link.target = '_blank';
    logo.parentNode.insertBefore(link, logo);
    link.appendChild(logo);
  }
});
</script>
```

# Day 1

# Welcome

## Objectives

- understand spatial data structures in R
- work operationally with vector and raster data
- move from tables to geometries
- align projections and study areas
- compute spatial relationships and summaries
- build analysis-ready outputs

## What This Workshop Is

- concept-driven first, package-specific second
- focused on using R as a GIS
- designed around two analytical routes:
  - points: simpler entry route
  - lines: more advanced route
- built from complete instructor workflows that become exercises

## What This Workshop Is Not

- not a full geodesy or cartography course
- not a survey of every spatial package in R
- not a client-specific data walkthrough

::: {.callout-caution}
We will spend less time on "what is the perfect projection?" and more time on "how do I inspect, transform, and use projections correctly in R?"
:::

## Learning Routes

:::: {.columns}
::: {.column width="50%"}
`Core route`

- choose one ecosystem:
  - `sf` + `stars`
  - `terra`
- choose one storyline:
  - points
  - lines
:::
::: {.column width="50%"}
`Advanced route`

- do both storylines in one ecosystem
- or do one storyline plus:
  - advanced GLM
  - KDE
  - both
:::
::::

## A Note About IDEs

- RStudio
- VS Code
- Positron

## Geocomputation Resources

- [Geocomputation with R](https://r.geocompx.org/)
- [SCI1031 notes](https://sci1031.github.io/sci1031/)
- [r-spatial ecosystem](https://github.com/r-spatial)

:::footer
[Geocomputation with R](https://r.geocompx.org/)
&nbsp; ***·*** &nbsp;
[r-spatial](https://github.com/r-spatial)
:::

# R As A GIS

## Why Use R As A GIS?

- one environment for import, cleaning, spatial processing, modeling, and reporting
- reproducible workflows instead of manual click sequences
- transparent transformations and analysis choices
- easy transition from exploration to automation

## Spatial Workflow Anatomy

1. import data
2. create or validate geometry
3. inspect CRS and transform if needed
4. restrict to a study area
5. measure, join, intersect, extract
6. summarise for analysis
7. model, map, and export

## Spatial Data At A Macro Level

:::: {.columns}
::: {.column width="50%"}
`Vector`

- discrete features
- geometry + attributes
- points, lines, polygons
- typical questions:
  - where?
  - how far?
  - what overlaps what?
:::
::: {.column width="50%"}
`Raster`

- gridded surfaces
- cells with values
- extent, resolution, CRS
- typical questions:
  - what value is here?
  - how does the surface vary?
  - what is the average over an area?
:::
::::

## Main Packages

:::: {.columns}
::: {.column width="33%"}
`sf`

- vector data
- simple features standard
- geometry-aware data frames
:::
::: {.column width="33%"}
`stars`

- raster and arrays
- good fit with `sf`
- convenient for plotting and extraction
:::
::: {.column width="33%"}
`terra`

- vector + raster in one package
- strong for raster operations
- consistent GIS-style workflow
:::
::::

## Supporting Packages

- `dplyr` for data manipulation
- `units` for explicit measurement handling in `sf`
- `mapview` for rapid interactive mapping
- `ggplot2` for polished static figures
- `MASS::kde2d()` for simple KDE examples

## How To Read The Rest Of This Deck

- slides are concept-driven first
- examples will often show an `sf/stars` view and a `terra` view
- the goal is not to memorize APIs
- the goal is to recognize the spatial task and know where the tools live

# Vectors

## Why Vectors First?

- vectors are usually more intuitive for beginners
- both workshop storylines start from tabular coordinates
- joins, intersections, and measurements are easier to explain here first
- raster integration makes more sense once the vector workflow is stable

## The Vector Data Model

- features have attributes and geometry
- geometry lives in a CRS
- geometry types matter:
  - points for observations
  - lines for trajectories
  - polygons for regions and habitats

## Two Vector Storylines

:::: {.columns}
::: {.column width="50%"}
`Points route`

- bird observations
- start from tabular lon/lat
- create points
- join to polygon context
:::
::: {.column width="50%"}
`Lines route`

- telemetry fixes
- start from lon/lat + timestamp
- create points
- rebuild tracks from ordered points
:::
::::

## Import Vector And Tabular Data

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
birds <- read.csv("bird_observations.csv")
ecodistricts <- st_read("ecodistricts.gpkg")
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
birds <- read.csv("bird_observations.csv")
ecodistricts <- vect("ecodistricts.gpkg")
```
:::
::::

- key idea: import attributes and import geometry are separate concerns

## Create Point Geometry From lon/lat

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
sites <- birds |>
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326,
           remove = FALSE)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
sites <- vect(birds,
              geom = c("longitude", "latitude"),
              crs = "EPSG:4326")
```
:::
::::

## Rebuild Lines From Ordered Points

- start from telemetry points
- sort by individual and timestamp
- group ordered coordinates into tracks
- line building is not a separate data source
- it is a geometry-construction step

::: {.notes}
This is an important pedagogic distinction: participants are not "reading lines", they are constructing them.
:::

## CRS: What We Need Operationally

- inspect current CRS
- know whether data are geographic or projected
- choose a working CRS for measurement and analysis
- transform all relevant layers consistently

## CRS API Lens

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
st_crs(sites)
sites_qc <- st_transform(sites, 32198)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
crs(sites)
sites_qc <- project(sites, "EPSG:32198")
```
:::
::::

## Restrict To The Study Area

- keep only the relevant geographic scope
- reduce clutter and computation
- make later extraction and summaries easier
- for vectors this is usually:
  - filter
  - crop
  - keep overlapping features

## Study-Area Restriction API Lens

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
sites <- st_filter(sites, east)
ecodistricts <- st_filter(ecodistricts, east)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
sites <- crop(sites, east)
ecodistricts <- crop(ecodistricts, east)
```
:::
::::

## Vector Measurements

- polygon area
- line length
- distance from points or tracks to coast
- measurement depends on CRS

::: {.callout-note}
`sf` often makes units explicit through `units`. `terra` usually returns numeric values whose interpretation depends on CRS and function behavior.
:::

## Measurement API Lens

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
ecodistricts$area <- st_area(ecodistricts)
sites$coast_distance <- st_distance(sites, coast)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
ecodistricts$area_km2 <- expanse(ecodistricts, unit = "km")
sites$coast_distance_km <- distance(sites, coast)[, 1] / 1000
```
:::
::::

## Vector Relationships

:::: {.columns}
::: {.column width="50%"}
`Join logic`

- attach polygon attributes to points
- ask which polygon contains each feature
- enrich observations with context
:::
::: {.column width="50%"}
`Intersection logic`

- create new geometries from overlap
- for line workflows, create habitat-attributed track segments
- this is not only an attribute join
:::
::::

## Relationship API Lens

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
sites <- st_join(sites, ecodistricts)
track_segments <- st_intersection(tracks, habitats)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
site_values <- extract(ecodistricts, sites)
track_segments <- intersect(tracks, habitats)
```
:::
::::

## Quick Vector Mapping

- `plot()` for fast sanity checks
- `mapview()` for interactive exploration
- the first map is often a QA/QC step, not a publication figure

## Exercise Block: Vector Foundations

- `M1`: import tables and vector layers, create geometry
- `M2`: harmonize CRS and restrict to study area
- `M3`: explore, measure, join, intersect

:::: {.columns}
::: {.column width="50%"}
`Points scripts`

- sections 1-7
:::
::: {.column width="50%"}
`Lines scripts`

- sections 1-9
:::
::::

## Route Choice After Vector Foundations

- most participants: continue with one storyline only
- recommended default:
  - points route for simpler progression
  - lines route for advanced participants
- strong participants can switch later to the other storyline

# Day 2

# Rasters

## The Raster Data Model

- a raster is a grid of cells with values
- key metadata:
  - extent
  - resolution
  - CRS
  - number of layers
- raster operations often modify the grid itself, not only attributes

## Why Rasters Need Their Own Section

- different mental model than vectors
- projection can imply resampling
- crop, mask, and extraction have different meanings than with vectors
- learners benefit from seeing this as a separate object system first

## Import And Inspect Rasters

:::: {.columns}
::: {.column width="50%"}
`stars`

```{r}
#| eval: false
temperature <- read_stars("temperature.tif")
plot(temperature)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
temperature <- rast("temperature.tif")
plot(temperature)
```
:::
::::

## What To Inspect Before Doing Anything Else

- CRS
- layer names
- extent
- resolution
- whether the raster already covers the study area

## Crop, Mask, Project

:::: {.columns}
::: {.column width="33%"}
`Crop`

- reduce extent
- keep a rectangular window or study boundary extent
:::
::: {.column width="33%"}
`Mask`

- keep values only inside a geometry
- remove cells outside a polygon
:::
::: {.column width="33%"}
`Project`

- change CRS
- usually implies resampling
- use carefully
:::
::::

## Raster Transformation API Lens

:::: {.columns}
::: {.column width="50%"}
`sf/stars`

```{r}
#| eval: false
bathymetry <- st_warp(bathymetry, crs = st_crs(tracks))
bathymetry <- st_crop(bathymetry, st_bbox(tracks))
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
bathymetry <- project(bathymetry, target_crs)
bathymetry <- crop(bathymetry, ext(tracks))
```
:::
::::

## Quick Raster Mapping

- use `plot()` to inspect values and spatial coverage
- use `mapview()` for interactive exploration where helpful
- use `geom_stars()` later when building a final static map

## Exercise Block: Raster Foundations

- `M4`: import and inspect raster data
- `M5`: crop, project, mask logic and prepare extraction

:::: {.columns}
::: {.column width="50%"}
`Points scripts`

- sections 8-11
:::
::: {.column width="50%"}
`Lines scripts`

- sections 10-13
:::
::::

# Integrated Workflows

## Why Integration Matters

- most analyses combine feature geometry with gridded context
- vectors define where
- rasters contribute environmental values
- analysis tables sit at the interface between the two

## Vector-Raster Extraction

- points: extract cell values at observation locations
- lines: extract summaries along tracks or segments
- segments: connect movement geometry to environmental context

## Extraction API Lens

:::: {.columns}
::: {.column width="50%"}
`sf/stars`

```{r}
#| eval: false
sites$tavg_january <- st_extract(temperature, sites)[[1]][, 1]
segment_bathymetry <- st_extract(bathymetry, track_segments, FUN = mean)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
temperature_values <- extract(temperature, sites)
segment_bathymetry <- extract(bathymetry, track_segments, fun = mean)
```
:::
::::

## Build Analysis Tables

- keep only the attributes needed for modeling
- drop geometry when the spatial step is complete
- summarise by polygon, track, or segment
- compute grouped metrics for later analysis

## Table-Building Verbs

- `st_drop_geometry()` or `as.data.frame()`
- `select()`
- `mutate()`
- `group_by()`
- `summarise()`
- `count()`
- `left_join()`

## Analysis Menu In This Workshop

:::: {.columns}
::: {.column width="50%"}
`Core`

- counts by ecodistrict
- grouped summaries
- final mapped outputs
:::
::: {.column width="50%"}
`Advanced`

- pseudo-absence GLM for points
- Gamma GLM on line segments
- KDE for points or telemetry fixes
:::
::::

## Simple Aggregated Modeling

- points route:
  - counts by ecodistrict
  - simple Poisson GLM
- lines route:
  - summaries by track or segment
  - model-ready tables from overlap and extraction

## Advanced Analysis: Pseudo-Absence GLM

- points workflow only
- sample pseudo-absences
- extract the same predictors as for presences
- fit a simple binomial GLM

## Advanced Analysis: Segment-Level GLM

- lines workflow only
- use habitat-attributed segments
- integrate segment length, environmental summaries, and habitat context
- fit a Gamma GLM on segment-level outcomes

## Advanced Analysis: KDE

- available in both storylines
- used to create a smooth surface of concentration or intensity
- useful for exploratory visualization and discussion

::: {.callout-note}
In the workshop scripts, KDE values are scaled to 0-1 for plotting convenience only. The raw density concept is unchanged.
:::

## Final Maps

- combine vector layers, raster context, and derived outputs
- distinguish quick QA/QC maps from communicative figures
- choose the simplest map that answers the question

## Export Outputs

- export vector outputs
- export summary tables
- keep derived objects for later reporting or modeling

:::: {.columns}
::: {.column width="50%"}
`sf/stars`

- `st_write()`
- `write.csv()`
:::
::: {.column width="50%"}
`terra`

- `writeVector()`
- `write.csv()`
:::
::::

## Exercise Architecture

1. `M1-M3`: vector foundations
2. `M4-M5`: raster foundations
3. `M6`: summaries and analysis tables
4. `M7-M8`: advanced options
5. `M9`: final map and export

## Script Split: Points

`Core beginner`

- sections 1-12
- import
- create points
- CRS
- study area
- measurements and joins
- raster extraction
- summaries

`Optional advanced`

- section 13: simple GLM
- sections 14-15: pseudo-absence GLM and KDE
- sections 16-17: final map and export

## Script Split: Lines

`Core advanced`

- sections 1-14
- import
- create points
- rebuild tracks
- CRS
- study area
- measurements
- habitat intersection
- raster extraction
- summaries

`Optional advanced`

- sections 15-16: segment GLM and KDE
- sections 17-18: final map and export

## Suggested Delivery Logic

- everyone sees the same concept deck
- everyone chooses one package ecosystem
- most participants complete either:
  - points core + one advanced option
  - lines core + one advanced option
- advanced participants may do both storylines in one ecosystem

## Extension: Workflow Structuring

- write small reusable functions
- group them into a package
- expose a workflow through Shiny when useful
- this is a conceptual endpoint for the workshop, not a participant coding task

## Takeaways

- think in spatial tasks, not only package names
- separate vector and raster mental models before integrating them
- use CRS deliberately
- distinguish joins from intersections
- move from spatial objects to analysis tables consciously
- use mapping throughout the workflow, not only at the end

## Next Step In The Hands-On

- choose your ecosystem:
  - `sf` + `stars`
  - `terra`
- choose your storyline:
  - points
  - lines
- begin with `M1`: import and geometry creation
