---
title: "R in Space"
author: "*inSileco*"
date: "2026-03-25"
from: markdown+emoji
---

```{=html}
<script>
document.addEventListener('DOMContentLoaded', function() {
  var logo = document.querySelector('.slide-logo');
  if (logo) {
    var link = document.createElement('a');
    link.href = 'https://www.insileco.io';
    link.target = '_blank';
    logo.parentNode.insertBefore(link, logo);
    link.appendChild(logo);
  }
});
</script>
```

<!-- Data & library setup -->
```{r}
#| message: false
#| warning: false
#| echo: false
source("setup_slide_data.R", local = knitr::knit_global())
```



# Welcome 

<!-- Go around the table + slide insileco? -->
<!-- What do you expect  -->

# Objectives

## What To Expect

:::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## What This Workshop Is

- a practical introduction to spatial workflows in R
- a workshop about using R operationally as a GIS
- focused on common manipulations you will reuse in real projects
:::
:::
::: {.column width="50%"}
::: {.callout-warning}
## What This Workshop Is Not

- not a full course in geodesy, cartography, or remote sensing
- not a survey of every spatial package or method
- not a theory-first geocomputation course
:::
:::
::::


## Resources

- [Visualisation et analyse de données spatiales sous R (FR)](https://sci1031.github.io/sci1031/)
- [Geocomputation with R](https://r.geocompx.org/)
- [Spatial Data Science](https://r-spatial.org/book/)

## Workshop Objective

By the end of the workshop, participants should be able to use R as a GIS, work operationally with common spatial data manipulations, and have the right tools in hand to keep learning.

## Our Focus

- import spatial data
- inspect geometry and CRS
- transform and subset data
- measure distance, area, and length
- combine layers through joins and overlays
- extract raster values to vector features
- build summaries, analyses, and maps

## Workshop Structure

:::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## Materials

- short concept-driven slides
- live examples in R
- reusable scripts and resources for later review
:::
:::
::: {.column width="50%"}
::: {.callout-tip}
## Exercises

- targeted hands-on exercises
- designed specifically for workshop participants
- based on the data shared ahead of the event
:::
:::
::::

::: {.callout icon=false}
The workshop alternates between concise instructional material and applied exercises, so participants can move directly from concepts to practice.
:::

# R as a GIS

<!-- Discuss why it makes sense to use R as a GIS -->

## R to build reproducible pipelines

![](img/flow.svg)

::: {.notes}
Could be python etc...
Similar as QGIS for C libraries
:::


## Spatial Data

<table class="spatial-objects-table">
  <thead>
    <tr>
      <th>Object type</th>
      <th>Description</th>
      <th><code>sf</code></th>
      <th><code>stars</code></th>
      <th><code>terra</code></th>
    </tr>
  </thead>
  <tbody>
    <tr class="family-row">
      <td>Vector</td>
      <td class="family-note"><em>Where are the features?</em></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td><img src="img/point.svg" class="obj-icon" alt="Point icon"> Points</td>
      <td>Discrete locations with coordinates and attributes.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr>
      <td><img src="img/line.svg" class="obj-icon" alt="Line icon"> Lines</td>
      <td>Linear features such as roads, rivers, or tracks.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr>
      <td><img src="img/polygon.svg" class="obj-icon" alt="Polygon icon"> Polygons</td>
      <td>Areas such as lakes, study zones, or administrative boundaries.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr class="family-row">
      <td>Raster</td>
      <td class="family-note"><em>What is the value across space?</em></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td><img src="img/grid.svg" class="obj-icon" alt="Raster grid icon"> Rasters</td>
      <td>Gridded surfaces with values stored in cells.</td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
  </tbody>
</table>

## Main Spatial R Packages

***New ecosystem***

:::: {.columns}
::: {.column width="33%"}

::: {.callout-tip icon=false}
### `sf` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/sf/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="sf documentation">&#127760;</a></span>

- *simple features* [standard](https://www.iso.org/standard/40114.html)
- vector data
- tidy friendly
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `stars` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/stars/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="stars documentation">&#127760;</a></span>

- spatiotemporal arrays
- raster and datacubes
- strong companion to `sf`
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `terra` <span class="pkg-link-wrap"><a href="https://rspatial.github.io/terra/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="terra documentation">&#127760;</a></span>

- vector and raster support
- strong raster tooling
- full GIS-style workflow
:::
:::
::::

***Old ecosystem***

:::: {.columns}
::: {.column width="33%"}

::: {.callout-important icon=false}
### `sp` <span class="pkg-link-wrap"><a href="https://cran.r-project.org/web/packages/sp/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="sp documentation">&#127760;</a></span>

- vector data
:::
:::
::: {.column width="33%"}
::: {.callout-important icon=false}
### `raster` <span class="pkg-link-wrap"><a href="https://cran.r-project.org/web/packages/raster/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="raster documentation">&#127760;</a></span>

- rasters
:::
:::
::: {.column width="33%"}
:::
::::

## Accompanying Packages

***Data manipulation and measurements***

:::: {.columns}
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `tidyverse` <span class="pkg-link-wrap"><a href="https://www.tidyverse.org/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tidyverse website">&#127760;</a></span>

- tabular workflows around spatial objects
- filtering, joining, grouping, summarising
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `tidyterra` <span class="pkg-link-wrap"><a href="https://dieghernan.github.io/tidyterra/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tidyterra documentation">&#127760;</a></span>

- tidy verbs and plotting helpers for `terra`
- a useful extension, not a replacement for `terra`
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `units` <span class="pkg-link-wrap"><a href="https://r-quantities.github.io/units/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="units documentation">&#127760;</a></span>

- explicit lengths, areas, and distances in `sf`
- contrasts `terra`, returns numeric measurements
:::
:::
::::

***Visualization***

:::: {.columns}
::: {.column width="33%"}
::: {.callout-note icon=false}
### `ggplot2` <span class="pkg-link-wrap"><a href="https://ggplot2.tidyverse.org/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="ggplot2 website">&#127760;</a></span>

- communication-quality static graphics
- flexible layered maps
:::
:::
::: {.column width="33%"}
::: {.callout-note icon=false}
### `mapview` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/mapview/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="mapview documentation">&#127760;</a></span>

- fast interactive exploration
- quick visual checks of spatial layers
:::
:::
::: {.column width="33%"}
::: {.callout-note icon=false}
### `tmap` <span class="pkg-link-wrap"><a href="https://r-tmap.github.io/tmap/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tmap website">&#127760;</a></span>

- thematic mapping workflows
- static and interactive map support
:::
:::
::::

::: {.callout-warning}
#### These are examples of particularly relevant accompanying packages. The spatial ecosystem in R is broad and active, and is not limited to the packages shown here.
:::

## A Note on IDEs

*What are IDEs? Integrated Development Environments*

:::: {.columns}
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `RStudio` <span class="pkg-link-wrap"><a href="https://posit.co/products/open-source/rstudio/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="RStudio website">&#127760;</a></span>

- long-standing default environment for R
- integrated console, editor, plots, files, and packages
- a strong starting point for most R users
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `VS Code` <span class="pkg-link-wrap"><a href="https://code.visualstudio.com/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="VS Code website">&#127760;</a></span>

- general-purpose editor with R support through extensions
- suited for mixed workflows across R, Python, Quarto, Git
- a good option for multi-language projects
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `Positron` <span class="pkg-link-wrap"><a href="https://positron.posit.co/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="Positron website">&#127760;</a></span>

- newer data-science IDE from Posit, VS Code-like
- combines scripting, notebooks, and IDE features
- worth watching as the ecosystem evolves
:::
:::
::::

::: {.callout-warning}
### The goal of this workshop is not to teach a specific IDE. Use the environment that lets you work comfortably with scripts, objects, plots, and spatial outputs.
:::

## A Note on Pipes 

::: {.callout-note}
### What are pipes?

Pipes pass the output of one step into the next, making multi-step workflows easier to read and write. We mention them explicitly because they are used throughout the workshop.
:::

<br>

:::: {.columns}

::: {.column width="50%"}

`magrittr` ***pipes:*** `%>%`

```R
CO2 %>%
  dplyr::filter(Type == "Quebec")
```
:::

::: {.column width="50%"}

***native pipes:*** `|>`

```R
CO2 |>
  dplyr::filter(Type == "Quebec")
```

:::

::::

## A Note on Code

***Choose Your Tool(s)***

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| eval: false
sf_object <- sf::function(...)
stars_object <- stars::function(...)
```

##### `terra`

```{r}
#| eval: false
terra_object <- terra::function(...)
```
:::

<br> 

::: {.callout-note}
### Navigating through the code in the presentation
The slides often show the same operation twice: once with `sf/stars`, and once with `terra`.

You do not need to follow both versions at once. Use the tabs to stay in the ecosystem you chose for the workshop.
:::



## A Note on Exercises

***Choose Your Route***

![](img/route_matrix.svg){width=82% fig-align="center"}

::: {.callout-note}
### Do both levels!
More advanced participants can complete both workflows, as there will be advanced workflows available with both points and lines. 
:::


# Vector Data in R

***sf*** & ***terra***

## Read Vector Data

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
aoi <- st_read("data/polygons.gpkg", layer = "study_area")
zones <- st_read("data/polygons.gpkg", layer = "zones")
aoi
zones
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(aoi)
head(zones)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
#| 
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2.5,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones),
  add = TRUE,
  col = c("#9ecae1aa", "#fdae6baa", "#a1d99baa", "#fdd0a2aa"),
  border = "#264653",
  lwd = 1.2
)
zone_centers <- st_coordinates(st_centroid(zones))
text(
  zone_centers[, 1],
  zone_centers[, 2],
  labels = zones$zone_id,
  cex = 0.9,
  font = 2,
  col = "#17323a"
)
```

:::
:::
<!-- end columns -->


##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
aoi <- vect("data/polygons.gpkg", layer = "study_area")
zones <- vect("data/polygons.gpkg", layer = "zones")
aoi
zones
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_aoi)
head(terra_zones)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2.5,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones,
  add = TRUE,
  col = c("#9ecae1aa", "#fdae6baa", "#a1d99baa", "#fdd0a2aa"),
  border = "#264653",
  lwd = 1.2
)
zone_centers <- terra::crds(terra::centroids(terra_zones))
text(
  zone_centers[, 1],
  zone_centers[, 2],
  labels = terra_zones$zone_id,
  cex = 0.9,
  font = 2,
  col = "#17323a"
)
```

:::
:::
<!-- end columns -->
:::

## Create Points From A Table

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
stations <- read.csv("data/points.csv") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(points_sf)
```
:::
::: {.column width="35%"}
<br>

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones),
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  st_geometry(points_sf),
  add = TRUE,
  pch = 21,
  bg = "#c44e5288",
  col = "#7f1d1d",
  cex = 1.1
)
```

:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
stations <- read.csv("data/points.csv") |>
  vect(geom = c("lon", "lat"), crs = "EPSG:4326")
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_points)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones,
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  terra_points,
  add = TRUE,
  pch = 21,
  bg = "#c44e5288",
  col = "#7f1d1d",
  cex = 1.1
)
```

:::
:::
<!-- end columns -->
:::


## Inspect Vector Data

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
st_geometry_type(zones)
st_crs(zones)$input
st_bbox(zones)
```
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
st_geometry_type(zones)
st_crs(zones)$input
st_bbox(zones)
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
geomtype(terra_zones)
crs(terra_zones, proj = TRUE)
ext(terra_zones)
```
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
geomtype(terra_zones)
crs(terra_zones, proj = TRUE)
ext(terra_zones)
```
:::

<br> 

::: {.callout-note}
### Spatial projections
CRS information tells you how coordinates are referenced on the Earth. When you need to inspect or identify a projection, [epsg.io](https://epsg.io/) is a useful lookup tool.
:::

## Transform CRS

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
zones_qc <- st_transform(zones, 32198)
points_qc <- st_transform(points_sf, 32198)
st_crs(points_qc)$input
st_bbox(points_qc)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
st_crs(points_sf_qc)$input
st_bbox(points_sf_qc)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi_qc),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones_qc),
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  st_geometry(points_sf_qc),
  add = TRUE,
  pch = 21,
  bg = "#c44e5288",
  col = "#7f1d1d",
  cex = 1.1
)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
zones_qc <- project(terra_zones, "EPSG:32198")
points_qc <- project(terra_points, "EPSG:32198")
crs(points_qc, proj = TRUE)
ext(points_qc)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
crs(terra_points_qc, proj = TRUE)
ext(terra_points_qc)
```
:::
::: {.column width="35%"}

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi_qc,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones_qc,
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  terra_points_qc,
  add = TRUE,
  pch = 21,
  bg = "#c44e5288",
  col = "#7f1d1d",
  cex = 1.1
)
```
:::
:::
<!-- end columns -->
:::

::: {.callout-warning}
## Measurements 
Use a suitable projected CRS when you need interpretable distances, lengths, or areas.
:::

## Export Vector Outputs

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
st_write(points_sf_qc, "outputs/points_projected.gpkg")
```

```{r}
#| message: false
#| warning: false
#| echo: false
sf_export <- tempfile(fileext = ".gpkg")
st_write(points_sf_qc, sf_export, quiet = TRUE)
data.frame(
  object = "points_sf_qc",
  file = basename(sf_export),
  exists = file.exists(sf_export),
  features = nrow(points_sf_qc)
)
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
writeVector(terra_points_qc, "outputs/points_projected.gpkg")
```

```{r}
#| message: false
#| warning: false
#| echo: false
terra_export <- tempfile(fileext = ".gpkg")
writeVector(terra_points_qc, terra_export, overwrite = TRUE)
data.frame(
  object = "terra_points_qc",
  file = basename(terra_export),
  exists = file.exists(terra_export),
  features = nrow(terra_points_qc)
)
```
:::

## Exercise 1 {.exercise-break .center}

### Vector Foundations

Import, inspect, transform, export

<br>

*Bonus: use R as a converter*

## Build Lines From Ordered Points

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
tracks <- points_sf |>
  dplyr::arrange(track_id, timestamp) |>
  dplyr::group_by(track_id) |>
  dplyr::summarise(do_union = FALSE) |>
  st_cast("LINESTRING")
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(tracks_sf)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones),
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  st_geometry(tracks_sf),
  add = TRUE,
  col = "#7f1d1d",
  lwd = 2.3
)
plot(
  st_geometry(points_sf),
  add = TRUE,
  pch = 16,
  col = "#7f1d1daa",
  cex = 0.65
)
```

:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
tracks_tbl <- points_tbl |>
  dplyr::arrange(track_id, timestamp) |>
  dplyr::summarise(
    wkt = paste0(
      "LINESTRING (",
      paste(paste(lon, lat), collapse = ", "),
      ")"
    ),
    .by = track_id
  ) |>
  terra::vect(geom = "wkt", crs = "EPSG:4326")
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_tracks)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones,
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  terra_tracks,
  add = TRUE,
  col = "#7f1d1d",
  lwd = 2.3
)
plot(
  terra_points,
  add = TRUE,
  pch = 16,
  col = "#7f1d1daa",
  cex = 0.65
)
```

:::
:::
<!-- end columns -->
:::

## Quick Static Mapping

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| eval: false
plot(st_geometry(aoi), col = "#f7fbfc", border = "#355c67")
plot(st_geometry(zones), add = TRUE, border = "#264653")
plot(st_geometry(stations), pch = 16, col = "#7f1d1d", add = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center
#| out-width: 60%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones),
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  st_geometry(points_sf),
  add = TRUE,
  pch = 16,
  col = "#7f1d1d",
  cex = 0.9
)
```

##### `terra`

```{r}
#| eval: false
plot(aoi, col = "#f7fbfc", border = "#355c67")
plot(zones, add = TRUE, border = "#264653")
plot(stations, pch = 16, col = "#7f1d1d", add = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center
#| out-width: 60%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones,
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  terra_points,
  add = TRUE,
  pch = 16,
  col = "#7f1d1d",
  cex = 0.9
)
```
:::

## Quick Interactive Mapping

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| eval: false
mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(stations, zcol = "category", layer.name = "Observations")
```

```{r}
#| message: false
#| warning: false
#| echo: false
mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(points_sf, zcol = "category", layer.name = "Observations")
```

##### `terra`

```{r}
#| eval: false
mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(stations, zcol = "category", layer.name = "Observations")
```

```{r}
#| message: false
#| warning: false
#| echo: false
mapview(terra_zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(terra_points, zcol = "category", layer.name = "Observations")
```
:::

## Export Interactive Maps

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| eval: false
m <- mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(stations, zcol = "category", layer.name = "Observations")

mapview::mapshot(m, file = "outputs/interactive_map.html")
```

```{r}
#| message: false
#| warning: false
#| echo: false
sf_map_file <- tempfile(fileext = ".html")
data.frame(
  object = "mapview sf map",
  file = basename(sf_map_file),
  format = "html"
)
```

##### `terra`

```{r}
#| eval: false
m <- mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(stations, zcol = "category", layer.name = "Observations")

mapview::mapshot(m, file = "outputs/interactive_map.html")
```

```{r}
#| message: false
#| warning: false
#| echo: false
terra_map_file <- tempfile(fileext = ".html")
data.frame(
  object = "mapview terra map",
  file = basename(terra_map_file),
  format = "html"
)
```
:::

<br>

::: {.callout-note}
### Share your maps!
Interactive `mapview` outputs can be saved as standalone HTML files and shared like any other web document.
:::

## Exercise 2 {.exercise-break .center}

### Vector Foundations on Your Data

Import, inspect, transform, export, visualize


## Subset To A Study Area

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
focus_area <- zones |>
  dplyr::filter(zone_id %in% c("NE", "SE")) |>
  dplyr::summarise()

points_focus <- st_filter(points_sf, focus_area)
zones_focus <- st_crop(zones, st_bbox(focus_area))
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(points_focus_sf)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(st_geometry(focus_area), add = TRUE, col = "#37abc833", border = "#37abc8", lwd = 2)
plot(st_geometry(points_sf), add = TRUE, pch = 16, col = "#9aa5ab88", cex = 0.7)
plot(st_geometry(points_focus_sf), add = TRUE, pch = 21, bg = "#c44e5288", col = "#7f1d1d", cex = 1.1)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
focus_area <- terra_zones[terra_zones$zone_id %in% c("NE", "SE")]
points_focus <- crop(terra_points, focus_area)
zones_focus <- crop(terra_zones, ext(focus_area))
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_points_focus)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(terra_focus_area, add = TRUE, col = "#37abc833", border = "#37abc8", lwd = 2)
plot(terra_points, add = TRUE, pch = 16, col = "#9aa5ab88", cex = 0.7)
plot(terra_points_focus, add = TRUE, pch = 21, bg = "#c44e5288", col = "#7f1d1d", cex = 1.1)
```
:::
:::
<!-- end columns -->
:::

## Measure Vector Features

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
zone_area <- units::set_units(st_area(zones_qc), "km^2")
track_length <- units::set_units(st_length(tracks_sf_qc), "km")
boundary_distance <- units::set_units(st_distance(points_sf_qc, st_boundary(aoi_qc)), "km")
zone_area
track_length
head(boundary_distance)
```

::: {.columns}
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
zone_area_sf
track_length_sf
```
:::
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
head(point_boundary_sf)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
zone_area <- expanse(terra_zones_qc, unit = "km")
track_length <- perim(terra_tracks_qc) / 1000
boundary_distance <- distance(terra_points_qc, as.lines(terra_aoi_qc))[, 1] / 1000
zone_area
track_length
head(boundary_distance)
```

::: {.columns}
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
zone_area_terra
track_length_terra
```
:::
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
head(point_boundary_terra)
```
:::
:::
<!-- end columns -->

:::

<br>

::: {.callout-note icon=false}
`sf` often returns unit-aware measurements. `terra` often returns numeric values whose meaning depends on the CRS and function used.
:::

## Join Attributes To Features

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
points_joined <- st_join(points_sf, zones[, c("zone_id", "zone_type")])
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(sf::st_drop_geometry(points_joined_sf))
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
zone_cols <- c(NW = "#2c7fb8", NE = "#f03b20", SW = "#31a354", SE = "#fdae61")
par(mar = c(0, 0, 0, 0))
plot(st_geometry(aoi), col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(st_geometry(zones), add = TRUE, col = paste0(zone_cols[zones$zone_id], "33"), border = "#264653", lwd = 1)
plot(st_geometry(points_joined_sf), add = TRUE, pch = 21, bg = zone_cols[points_joined_sf$zone_id], col = "#17323a", cex = 1.1)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
point_values <- extract(terra_zones, terra_points)
points_joined <- cbind(terra_points, point_values[, c("zone_id", "zone_type")])
```

::: {.columns}
::: {.column width="68%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_points_joined)
```
:::
::: {.column width="32%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
zone_cols <- c(NW = "#2c7fb8", NE = "#f03b20", SW = "#31a354", SE = "#fdae61")
par(mar = c(0, 0, 0, 0))
plot(terra_aoi, col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(terra_zones, add = TRUE, col = paste0(zone_cols[terra_zones$zone_id], "33"), border = "#264653", lwd = 1)
plot(terra_points_joined, add = TRUE, pch = 21, bg = zone_cols[terra_points_joined$zone_id], col = "#17323a", cex = 1.1)
```
:::
:::
<!-- end columns -->
:::

## Intersect Geometries

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
track_segments <- st_intersection(tracks_sf_qc, zones_qc)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(sf::st_drop_geometry(track_segments_sf))
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
zone_cols <- c(NW = "#2c7fb8", NE = "#f03b20", SW = "#31a354", SE = "#fdae61")
par(mar = c(0, 0, 0, 0))
plot(st_geometry(aoi_qc), col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(st_geometry(zones_qc), add = TRUE, col = paste0(zone_cols[zones_qc$zone_id], "22"), border = "#264653", lwd = 1)
plot(st_geometry(track_segments_sf), add = TRUE, col = zone_cols[track_segments_sf$zone_id], lwd = 2.5)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
track_segments <- intersect(terra_tracks_qc, terra_zones_qc)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_track_segments)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
zone_cols <- c(NW = "#2c7fb8", NE = "#f03b20", SW = "#31a354", SE = "#fdae61")
par(mar = c(0, 0, 0, 0))
plot(terra_aoi_qc, col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(terra_zones_qc, add = TRUE, col = paste0(zone_cols[terra_zones_qc$zone_id], "22"), border = "#264653", lwd = 1)
plot(terra_track_segments, add = TRUE, col = zone_cols[terra_track_segments$zone_id], lwd = 2.5)
```
:::
:::
<!-- end columns -->
:::

::: {.callout-note icon=false}
Use a join when you want to attach attributes. Use an intersection when you want overlap to create new geometries.
:::

## Buffers

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
point_buffers <- st_buffer(points_sf_qc, dist = 10000)
st_geometry_type(point_buffers)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
st_geometry_type(point_buffers_sf)[1]
head(sf::st_drop_geometry(point_buffers_sf))
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(st_geometry(aoi_qc), col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(st_geometry(zones_qc), add = TRUE, col = "#9ecae122", border = "#264653", lwd = 1)
plot(st_geometry(point_buffers_sf), add = TRUE, col = "#37abc833", border = "#37abc8", lwd = 1)
plot(st_geometry(points_sf_qc), add = TRUE, pch = 16, col = "#7f1d1d", cex = 0.7)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
point_buffers <- buffer(terra_points_qc, width = 10000)
geomtype(point_buffers)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
geomtype(terra_point_buffers)
head(terra_point_buffers)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(terra_aoi_qc, col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(terra_zones_qc, add = TRUE, col = "#9ecae122", border = "#264653", lwd = 1)
plot(terra_point_buffers, add = TRUE, col = "#37abc833", border = "#37abc8", lwd = 1)
plot(terra_points_qc, add = TRUE, pch = 16, col = "#7f1d1d", cex = 0.7)
```
:::
:::
<!-- end columns -->
:::

::: {.callout-note icon=false}
Buffers create new geometries at a fixed distance around features, so they are usually done in a projected CRS with interpretable distance units.
:::

## Exercise 3 {.exercise-break .center}

### Vector Operations

Subset, measure, join, intersect, buffer


# Raster Data in R

***stars*** & ***terra***


## Read Raster Data

::: {.panel-tabset .code-tabs}
##### `stars`

```{r}
#| message: false
#| warning: false
#| eval: false
surface <- read_stars("data/surface.tif")
surface
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
surface_stars
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
plot(surface_stars, reset = FALSE, axes = FALSE, main = "")
plot(st_geometry(aoi), add = TRUE, border = "#355c67", lwd = 2)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
surface <- rast("data/surface.tif")
surface
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
surface_terra
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(surface_terra, axes = FALSE, main = "")
plot(terra_aoi, add = TRUE, border = "#355c67", lwd = 2)
```
:::
:::
<!-- end columns -->
:::

## Inspect Raster Data

::: {.panel-tabset .code-tabs}
##### `stars`

```{r}
#| message: false
#| warning: false
#| eval: false
st_dimensions(surface)
st_bbox(surface)
st_crs(surface)$input
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
st_dimensions(surface_stars)
st_bbox(surface_stars)
st_crs(surface_stars)$input
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
dim(surface)
res(surface)
ext(surface)
crs(surface, proj = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
dim(surface_terra)
res(surface_terra)
ext(surface_terra)
crs(surface_terra, proj = TRUE)
```
:::

## Export Raster Outputs

::: {.panel-tabset .code-tabs}
##### `stars`

```{r}
#| message: false
#| warning: false
#| eval: false
write_stars(surface_mask, "outputs/surface_focus.tif")
```

```{r}
#| message: false
#| warning: false
#| echo: false
stars_export <- tempfile(fileext = ".tif")
write_stars(surface_stars_mask, stars_export)
data.frame(
  object = "surface_stars_mask",
  file = basename(stars_export),
  exists = file.exists(stars_export)
)
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
writeRaster(surface_mask, "outputs/surface_focus.tif", overwrite = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
terra_raster_export <- tempfile(fileext = ".tif")
writeRaster(surface_terra_mask, terra_raster_export, overwrite = TRUE)
data.frame(
  object = "surface_terra_mask",
  file = basename(terra_raster_export),
  exists = file.exists(terra_raster_export)
)
```
:::

## Quick Static Raster Plotting

::: {.panel-tabset .code-tabs}
##### `stars`

```{r}
#| eval: false
plot(surface, col = viridis::viridis(100))
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center
#| out-width: 60%
plot(surface_stars, reset = FALSE, axes = FALSE, main = "", col = viridis::viridis(100))
```

##### `terra`

```{r}
#| eval: false
plot(surface)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center
#| out-width: 65%
par(mar = c(0, 0, 0, 0))
plot(surface_terra, axes = FALSE, main = "")
plot(terra_aoi, add = TRUE, border = "#355c67", lwd = 2)
```
:::

## Quick Interactive Raster Mapping

::: {.panel-tabset .code-tabs}
##### `stars`

```{r}
#| eval: false
mapview(surface) + mapview(aoi)
```

```{r}
#| message: false
#| warning: false
#| echo: false
mapview(surface_stars) + mapview(aoi)
```

##### `terra`

```{r}
#| eval: false
mapview(surface) + mapview(aoi)
```

```{r}
#| message: false
#| warning: false
#| echo: false
mapview(surface_terra) + mapview(terra_aoi)
```
:::


## Exercise 4 {.exercise-break .center}

### Raster Foundations

Import, inspect, transform, export, visualize


## Crop, Mask, Project

::: {.panel-tabset .code-tabs}
##### `stars`

```{r}
#| message: false
#| warning: false
#| eval: false
crop_zone <- zones |> dplyr::filter(zone_id == "NE")
buffer_one <- st_transform(point_buffers[1, ], st_crs(surface))
surface_qc <- st_warp(surface, crs = st_crs(aoi_qc))
surface_crop <- st_crop(surface, st_bbox(crop_zone))
surface_mask <- surface[buffer_one]
```

::: {.columns}
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| out-width: 100%
plot(surface_stars_qc, reset = FALSE, axes = FALSE, main = "Project")
plot(st_geometry(aoi_qc), add = TRUE, border = "#355c67", lwd = 2)
```
:::
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| out-width: 100%
plot(surface_stars_crop_demo, reset = FALSE, axes = FALSE, main = "Crop")
plot(st_geometry(crop_zone_sf), add = TRUE, border = "#355c67", lwd = 2)
```
:::
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| out-width: 100%
plot(surface_stars_mask_demo, reset = FALSE, axes = FALSE, main = "Mask")
plot(st_geometry(single_buffer_sf), add = TRUE, border = "#355c67", lwd = 1.5)
```
:::
:::

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
crop_zone <- zones[zones$zone_id == "NE"]
buffer_one <- project(point_buffers[1], crs(surface))
surface_qc <- project(surface, "EPSG:32198")
surface_crop <- crop(surface, crop_zone)
surface_mask <- mask(surface, buffer_one)
```

::: {.columns}
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| out-width: 100%
par(mar = c(0, 0, 1, 0))
plot(surface_terra_qc, axes = FALSE, main = "Project")
plot(terra_aoi_qc, add = TRUE, border = "#355c67", lwd = 2)
```
:::
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| out-width: 100%
par(mar = c(0, 0, 1, 0))
plot(surface_terra_crop_demo, axes = FALSE, main = "Crop")
plot(crop_zone_terra, add = TRUE, border = "#355c67", lwd = 2)
```
:::
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4
#| out-width: 100%
par(mar = c(0, 0, 1, 0))
plot(surface_terra_mask_demo, axes = FALSE, main = "Mask")
plot(single_buffer_terra, add = TRUE, border = "#355c67", lwd = 1.2)
```
:::
:::
:::

::: {.callout-note}
### Raster manipulations
Here the three operations are shown independently: `project` changes CRS, `crop` uses one zone as the target extent, and `mask` keeps cells only inside a single buffer geometry.
:::

## Resampling

::: {.panel-tabset .code-tabs}
##### `stars`

```{r}
#| message: false
#| warning: false
#| eval: false
template <- st_as_stars(st_bbox(surface), dx = 0.05, dy = 0.05)
st_crs(template) <- st_crs(surface)
surface_near <- st_warp(surface, dest = template, method = "near", use_gdal = TRUE)
surface_bilinear <- st_warp(surface, dest = template, method = "bilinear", use_gdal = TRUE)
```

::: {.columns}
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| out-width: 100%
plot(surface_stars, reset = FALSE, axes = FALSE, main = "Original")
```
:::
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| out-width: 100%
plot(surface_stars_near, reset = FALSE, axes = FALSE, main = "Near")
```
:::
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| out-width: 100%
plot(surface_stars_bilinear, reset = FALSE, axes = FALSE, main = "Bilinear")
```
:::
:::

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
template <- rast(ext(surface), resolution = 0.05, crs = crs(surface))
surface_near <- resample(surface, template, method = "near")
surface_bilinear <- resample(surface, template, method = "bilinear")
```

::: {.columns}
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| out-width: 100%
par(mar = c(0, 0, 1, 0))
plot(surface_terra, axes = FALSE, main = "Original")
```
:::
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| out-width: 100%
par(mar = c(0, 0, 1, 0))
plot(surface_terra_near, axes = FALSE, main = "Near")
```
:::
::: {.column width="33%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 5
#| fig-height: 4.5
#| out-width: 100%
par(mar = c(0, 0, 1, 0))
plot(surface_terra_bilinear, axes = FALSE, main = "Bilinear")
```
:::
:::
:::

::: {.callout-note}
### Resampling changes the cell grid
`Near` is common for categorical rasters, while `bilinear` is common for continuous surfaces.
:::

## Multiple Raster Layers

::: {.panel-tabset .code-tabs}
##### `stars`

```{r}
#| message: false
#| warning: false
#| eval: false
surface2 <- surface * 1.1
names(surface2) <- "surface2"
surface_stack <- c(surface, surface2)
surface_mean <- st_apply(surface_stack, MARGIN = c("x", "y"), FUN = mean)
```

::: {.columns}
::: {.column width="45%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
names(surface_stars_stack)
length(names(surface_stars_stack))
```
:::
::: {.column width="55%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5
#| out-width: 70%
plot(surface_stars_mean, reset = FALSE, axes = FALSE, main = "Cell-wise mean")
```
:::
:::

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
surface2 <- surface * 1.1
names(surface2) <- "surface2"
surface_stack <- c(surface, surface2)
surface_mean <- app(surface_stack, mean)
```

::: {.columns}
::: {.column width="45%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
names(surface_terra_stack)
nlyr(surface_terra_stack)
```
:::
::: {.column width="55%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5
#| out-width: 70%
par(mar = c(0, 0, 1, 0))
plot(surface_terra_mean, axes = FALSE, main = "Cell-wise mean")
```
:::
:::
:::

::: {.callout-note}
### Multi-layer rasters
Can represent bands, variables, or time steps. Cell-wise functions combine information across layers.
:::


## Exercise 5 {.exercise-break .center}

### Raster Operations

Project, crop, mask, resample, layer operations



# Integrated Workflows

## Working with Vectors & Rasters

![](img/integration_workflow.svg){width=92% fig-align="center"}


## Extract Raster Values To Points

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
points_surface <- st_extract(surface, points_joined)
names(points_surface)[1] <- "surface_value"
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(sf::st_drop_geometry(points_surface_sf))
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
plot(surface_stars, reset = FALSE, axes = FALSE, main = "")
plot(st_geometry(points_surface_sf), add = TRUE, pch = 21, bg = hcl.colors(nrow(points_surface_sf), "YlOrRd")[rank(points_surface_sf$surface_value)], col = "#17323a", cex = 1)
```
:::
:::

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
point_vals <- extract(surface, points_joined)
points_surface <- cbind(points_joined, point_vals[, "surface_value", drop = FALSE])
```

::: {.columns}
::: {.column width="60%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(as.data.frame(terra_points_surface)[, c("obs_id", "zone_id", "zone_type", "surface_value")])
```
:::
::: {.column width="40%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(surface_terra, axes = FALSE, main = "")
plot(terra_points_surface, add = TRUE, pch = 21, bg = hcl.colors(nrow(terra_points_surface), "YlOrRd")[rank(terra_points_surface$surface_value)], col = "#17323a", cex = 1)
```
:::
:::
:::

## Extract Raster Values To Lines

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
tracks_surface <- st_extract(surface, tracks, FUN = function(x) mean(x, na.rm = TRUE))
```

::: {.columns}
::: {.column width="60%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(sf::st_drop_geometry(tracks_surface_sf))
```
:::
::: {.column width="40%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
line_cols <- hcl.colors(nrow(tracks_surface_sf), "TealGrn")[rank(tracks_surface_sf$surface_value)]
plot(surface_stars, reset = FALSE, axes = FALSE, main = "")
plot(st_geometry(tracks_surface_sf), add = TRUE, col = line_cols, lwd = 2.5)
```
:::
:::

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
track_vals <- extract(surface, tracks, fun = mean, na.rm = TRUE)
tracks_surface <- cbind(tracks, track_vals[, "surface_value", drop = FALSE])
```

::: {.columns}
::: {.column width="60%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(as.data.frame(terra_tracks_surface)[, c("track_id", "surface_value")])
```
:::
::: {.column width="40%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
line_cols <- hcl.colors(nrow(terra_tracks_surface), "TealGrn")[rank(terra_tracks_surface$surface_value)]
par(mar = c(0, 0, 0, 0))
plot(surface_terra, axes = FALSE, main = "")
plot(terra_tracks_surface, add = TRUE, col = line_cols, lwd = 2.5)
```
:::
:::
:::

## Extract Raster Values To Polygons

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
zones_surface <- st_extract(surface, zones, FUN = function(x) mean(x, na.rm = TRUE))
```

::: {.columns}
::: {.column width="60%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(sf::st_drop_geometry(zones_surface_sf))
```
:::
::: {.column width="40%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
poly_cols <- hcl.colors(nrow(zones_surface_sf), "Sunset")[rank(zones_surface_sf$surface_value)]
plot(st_geometry(zones_surface_sf), col = poly_cols, border = "#264653", lwd = 1, axes = FALSE, main = "")
```
:::
:::

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
zone_vals <- extract(surface, zones, fun = mean, na.rm = TRUE)
zones_surface <- cbind(zones, zone_vals[, "surface_value", drop = FALSE])
```

::: {.columns}
::: {.column width="60%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(as.data.frame(terra_zones_surface)[, c("zone_id", "zone_type", "surface_value")])
```
:::
::: {.column width="40%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
poly_cols <- hcl.colors(nrow(terra_zones_surface), "Sunset")[rank(terra_zones_surface$surface_value)]
par(mar = c(0, 0, 0, 0))
plot(terra_zones_surface, col = poly_cols, border = "#264653", lwd = 1, axes = FALSE, main = "")
```
:::
:::
:::

## Summarise Extracted Values

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
point_zone_summary <- points_surface |>
  st_drop_geometry() |>
  group_by(zone_id, zone_type) |>
  summarise(
    point_n = n(),
    mean_surface = mean(surface_value, na.rm = TRUE),
    mean_value = mean(value, na.rm = TRUE)
  )
```

```{r}
#| message: false
#| warning: false
#| echo: false
point_zone_summary_sf
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
point_zone_summary <- as.data.frame(points_surface) |>
  group_by(zone_id, zone_type) |>
  summarise(
    point_n = n(),
    mean_surface = mean(surface_value, na.rm = TRUE),
    mean_value = mean(value, na.rm = TRUE)
  )
```

```{r}
#| message: false
#| warning: false
#| echo: false
point_zone_summary_terra
```
:::

## Build Analysis Tables

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
analysis_table <- point_zone_summary |>
  left_join(zone_area, by = "zone_id") |>
  select(zone_id, zone_type, point_n, mean_surface, mean_value, area_km2)
```

```{r}
#| message: false
#| warning: false
#| echo: false
analysis_table_sf
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
analysis_table <- point_zone_summary |>
  left_join(zone_area, by = "zone_id") |>
  select(zone_id, zone_type, point_n, mean_surface, mean_value, area_km2)
```

```{r}
#| message: false
#| warning: false
#| echo: false
analysis_table_terra
```
:::

## Exercise 6 {.exercise-break .center}

### Integrated Workflows

Extract, summarise, build analysis tables

# Spatial Analytics

## Spatial Analytics

- spatial analytics is a very broad field
- this workshop does not aim to cover that full breadth
- our emphasis is on getting spatial data ready for analysis
- we use simple GLMs and KDEs as illustrative examples of what can follow

::: {.callout-note}
### The point here is not the analyses themselves. The point is the workflow that gets you from spatial data to analysis-ready inputs.
:::

## Prepare Data For Analysis

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
analysis_data <- points_surface |>
  st_drop_geometry() |>
  select(obs_id, track_id, zone_id, zone_type, category, value, surface_value) |>
  filter(!is.na(zone_id), !is.na(surface_value))
```

```{r}
#| message: false
#| warning: false
#| echo: false
head(analysis_data_sf)
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
analysis_data <- as.data.frame(points_surface) |>
  select(obs_id, track_id, zone_id, zone_type, category, value, surface_value) |>
  filter(!is.na(zone_id), !is.na(surface_value))
```

```{r}
#| message: false
#| warning: false
#| echo: false
head(analysis_data_terra)
```
:::

## A Simple GLM

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
glm_fit <- glm(
  value ~ surface_value + zone_type,
  family = poisson(),
  data = analysis_data
)
summary(glm_fit)
```

```{r}
#| message: false
#| warning: false
#| echo: false
glm_coef_sf
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
glm_fit <- glm(
  value ~ surface_value + zone_type,
  family = poisson(),
  data = analysis_data
)
summary(glm_fit)
```

```{r}
#| message: false
#| warning: false
#| echo: false
glm_coef_terra
```
:::

## Bonus: Advanced GLM

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
pseudo_points <- st_sample(aoi_qc, size = nrow(points_qc), exact = TRUE) |>
  st_as_sf() |>
  st_join(zones_qc[, c("zone_id", "zone_type")])

pseudo_points$surface_value <- st_extract(surface_qc, pseudo_points)[[1]]
pseudo_points$presence <- 0
points_qc$presence <- 1

pa_data <- bind_rows(points_qc, pseudo_points) |>
  st_drop_geometry() |>
  filter(!is.na(zone_id), !is.na(surface_value))

glm_pa <- glm(presence ~ surface_value + zone_type, family = binomial(), data = pa_data)
```

::: {.columns}
::: {.column width="55%"}
```{r}
#| message: false
#| warning: false
#| echo: false
glm_pa_coef_sf
```
:::
::: {.column width="45%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 70%
par(mar = c(0, 0, 0, 0))
plot(st_geometry(aoi_qc), col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(st_geometry(pseudo_points_sf), add = TRUE, pch = 1, col = "#7aa6c2", cex = 0.9)
plot(st_geometry(presence_points_sf), add = TRUE, pch = 16, col = "#7f1d1d", cex = 0.8)
```
:::
:::

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
pseudo_points <- spatSample(aoi_qc, size = nrow(points_qc), method = "random", as.points = TRUE)
pseudo_zone <- extract(zones_qc, pseudo_points)
pseudo_surface <- extract(surface_qc, pseudo_points)
pseudo_points <- cbind(pseudo_points, pseudo_zone[, c("zone_id", "zone_type")], pseudo_surface[, "surface_value", drop = FALSE])
pseudo_points$presence <- 0
points_qc$presence <- 1

pa_data <- bind_rows(as.data.frame(points_qc), as.data.frame(pseudo_points)) |>
  filter(!is.na(zone_id), !is.na(surface_value))

glm_pa <- glm(presence ~ surface_value + zone_type, family = binomial(), data = pa_data)
```

::: {.columns}
::: {.column width="55%"}
```{r}
#| message: false
#| warning: false
#| echo: false
glm_pa_coef_terra
```
:::
::: {.column width="45%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 70%
par(mar = c(0, 0, 0, 0))
plot(terra_aoi_qc, col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(pseudo_points_terra, add = TRUE, pch = 1, col = "#7aa6c2", cex = 0.9)
plot(presence_points_terra, add = TRUE, pch = 16, col = "#7f1d1d", cex = 0.8)
```
:::
:::
:::

## Exercise 7 {.exercise-break .center}

### Spatial Analytics

Prepare data, fit simple GLMs

<br>

*Bonus: generate pseudo-absences for your GLM with points data*

## Kernel Density Estimation (KDE) 

***Use as an exploratory tools***

- useful for point patterns
- produces a smooth intensity surface
- helps visualize concentration or hotspots
- often used for exploration rather than formal inference

## Kernel Density Estimation (KDE) 

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
pts_xy <- st_coordinates(points_qc)
bbox <- st_bbox(aoi_qc)
points_kde <- MASS::kde2d(
  pts_xy[, 1], pts_xy[, 2],
  n = 80,
  h = c(10000, 10000),
  lims = c(bbox["xmin"], bbox["xmax"], bbox["ymin"], bbox["ymax"])
)
```

```{r}
#| message: false
#| warning: false
#| echo: false
list(
  x_head = round(head(points_kde_sf_raw$x), 1),
  y_head = round(head(points_kde_sf_raw$y), 1),
  z_dim = dim(points_kde_sf_raw$z)
)
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
pts_xy <- crds(points_qc)
bbox <- ext(aoi_qc)
points_kde <- MASS::kde2d(
  pts_xy[, 1], pts_xy[, 2],
  n = 80,
  h = c(10000, 10000),
  lims = c(bbox$xmin, bbox$xmax, bbox$ymin, bbox$ymax)
)
```

```{r}
#| message: false
#| warning: false
#| echo: false
list(
  x_head = round(head(points_kde_terra_raw$x), 1),
  y_head = round(head(points_kde_terra_raw$y), 1),
  z_dim = dim(points_kde_terra_raw$z)
)
```
:::

## Kernel Density Estimation (KDE) 

::: {.panel-tabset .code-tabs}
##### `sf/stars`

```{r}
#| message: false
#| warning: false
#| eval: false
points_kde <- st_as_stars(
  list(kde = points_kde$z),
  dimensions = st_dimensions(x = points_kde$x, y = points_kde$y)
) |>
  st_set_crs(32198)

points_kde <- points_kde[aoi_qc]
points_kde_plot <- points_kde
points_kde_plot[[1]] <- points_kde_plot[[1]] / max(points_kde_plot[[1]], na.rm = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 50%
#| fig-align: center
kde_breaks <- seq(0, 1, by = 0.1)
kde_cols <- viridis::viridis(length(kde_breaks) - 1)
plot(st_geometry(aoi_qc), border = "#355c67", col = "#f7fbfc")
plot(points_kde_sf_plot, breaks = kde_breaks, col = kde_cols, add = TRUE)
plot(st_geometry(points_sf_qc), pch = 16, col = "#7f1d1d", cex = 0.55, add = TRUE)
plot(st_geometry(aoi_qc), border = "#355c67", col = NA, add = TRUE)
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
points_kde <- rast(
  t(points_kde$z)[nrow(t(points_kde$z)):1, ],
  extent = ext(min(points_kde$x), max(points_kde$x), min(points_kde$y), max(points_kde$y)),
  crs = "EPSG:32198"
)

points_kde <- mask(points_kde, aoi_qc)
points_kde_plot <- points_kde / global(points_kde, "max", na.rm = TRUE)[1, 1]
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 50%
#| fig-align: center
kde_breaks <- seq(0, 1, by = 0.1)
kde_cols <- viridis::viridis(length(kde_breaks) - 1)
plot(terra_aoi_qc, border = "#355c67", col = NA)
plot(points_kde_terra_plot, breaks = kde_breaks, col = kde_cols, add = TRUE)
points(terra_points_qc, pch = 16, cex = 0.55, col = "#7f1d1d")
plot(terra_aoi_qc, border = "#355c67", col = NA, add = TRUE)
```
:::

## Exercise 8 {.exercise-break .center}

### Spatial Analytics

Fit KDEs and explore surfaces

# Advanced Mapping

## Advanced Mapping

:::: {.columns}
::: {.column width="50%"}
`Quick exploration`

- `plot()`
- `mapview()`
- fast QA/QC and data inspection
:::
::: {.column width="50%"}
`Communication`

- `ggplot2`
- `tmap`
- more deliberate legends, palettes, labels, and layout
:::
::::

::: {.callout-note icon=false}
### Use exploratory maps to inspect data quickly, then move to more deliberate mapping tools when the goal is communication or sharing.
:::

## Advanced Mapping with `ggplot2`

`ggplot2` works especially well for polished static maps, particularly with `sf` and `stars` objects.

::: {.columns}
::: {.column width="50%"}
```{r}
#| eval: false
gg <- ggplot() +
  geom_stars(
    data = points_kde_sf_plot
  ) +
  scale_fill_viridis_c(name = "KDE") +
  geom_sf(
    data = zones_qc, 
    fill = NA, 
    color = "#355c67"
  ) +
  geom_sf(
    data = aoi_qc, 
    fill = NA, 
    color = "#17313b"
  ) +
  geom_sf(
    data = points_sf_qc, 
    aes(color = category), 
    size = 1.7
   ) +
  coord_sf(expand = FALSE) +
  theme_minimal()
gg
```
:::
::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 9
#| out-width: 100%
advanced_map_gg
```
:::
:::

<!-- end columns -->

## Advanced Mapping with `tmap`

`tmap` is built for thematic mapping and works well when you want a more map-focused grammar than `ggplot2`.

::: {.columns}
::: {.column}
```{r}
#| eval: false
tm <- tm_shape(points_kde_sf_plot) +
  tm_raster(col.scale = tm_scale_continuous(
    values = "viridis"
  )) +
  tm_shape(zones_qc) +
  tm_borders(col = "#355c67") +
  tm_shape(points_sf_qc) +
  tm_symbols(
    fill = "category", 
    size = 0.5
  ) +
  tm_shape(aoi_qc) +
  tm_borders(col = "#17313b") +
  tm_layout(
    frame = FALSE, 
    legend.outside = TRUE
  )
tm
```
:::
::: {.column}
```{r}
#| echo: false
#| fig-width: 9
#| fig-height: 9
#| out-width: 100%
tm
```
:::
:::
<!-- end columns -->



## Advanced Mapping with `tmap`

One advantage of `tmap` is that the same map logic can often be used for static output or for an interactive shared map.

```{r}
#| eval: false
tm_html <- tmap_leaflet(tm)
tm_html
```
```{r}
#| echo: false
#| warning: false
advanced_map_tmap_view
```

## Export And Share Maps

::: {.panel-tabset .code-tabs}
## `ggplot2`

```{r}
#| eval: false
ggsave(
  filename = "outputs/advanced_map.png",
  plot = gg,
  width = 8,
  height = 6,
  dpi = 300
)

ggsave(
  filename = "outputs/advanced_map.pdf",
  plot = gg,
  width = 8,
  height = 6
)
```

```{r}
#| echo: false
data.frame(
  output = c("presentation figure", "publication-ready figure"),
  file = c("advanced_map.png", "advanced_map.pdf"),
  tool = c("ggsave(..., dpi = 300)", "ggsave(...)")
)
```

## `tmap`

```{r}
#| eval: false
tmap_save(
  tm = tm,
  filename = "outputs/advanced_map.png",
  width = 8,
  height = 6,
  dpi = 300
)

tmap_mode("view")
tmap_save(
  tm = tm,
  filename = "outputs/advanced_map.html"
)
tmap_mode("plot")
```

```{r}
#| echo: false
data.frame(
  output = c("static thematic map", "interactive shared map"),
  file = c("advanced_map.png", "advanced_map.html"),
  tool = c("tmap_save(...)", "tmap_mode(\"view\") + tmap_save(...)")
)
```
:::

## Exercise 9 {.exercise-break .center}

### Advanced Mapping

Build, export, and share maps

## Takeaways

- vector and raster are different data models
- `sf`, `stars`, and `terra` cover most common spatial workflows in R
- CRS decisions affect measurements and overlays
- joins, intersections, crops, masks, and extraction are core operations
- spatial analysis in R is often a bridge between GIS tasks and standard statistical workflows
- treat R as both an analysis environment and a GIS

## Closing

- start with clear spatial objects and clear CRS
- inspect and map early, then analyse
- move from spatial objects to clean analysis tables
- keep building reusable workflows as your analyses grow
