---
title: "R in Space"
author: "*inSileco*"
date: "2026-03-25"
from: markdown+emoji
---

```{=html}
<script>
document.addEventListener('DOMContentLoaded', function() {
  var logo = document.querySelector('.slide-logo');
  if (logo) {
    var link = document.createElement('a');
    link.href = 'https://www.insileco.io';
    link.target = '_blank';
    logo.parentNode.insertBefore(link, logo);
    link.appendChild(logo);
  }
});
</script>
```

<!-- Data & library setup -->
```{r}
#| message: false
#| warning: false
#| echo: false
library(sf)
library(terra)
library(stars)
library(mapview)
aoi <- st_read("data/polygons.gpkg", layer = "study_area", quiet = TRUE)
zones <- st_read("data/polygons.gpkg", layer = "zones", quiet = TRUE)
terra_aoi <- terra::vect("data/polygons.gpkg", layer = "study_area")
terra_zones <- terra::vect("data/polygons.gpkg", layer = "zones")
points_tbl <- read.csv("data/points.csv")
points_sf <- st_as_sf(points_tbl, coords = c("lon", "lat"), crs = 4326, remove = FALSE)
terra_points <- terra::vect(points_tbl, geom = c("lon", "lat"), crs = "EPSG:4326")
tracks_sf <- points_sf |>
  dplyr::arrange(track_id, timestamp) |>
  dplyr::group_by(track_id) |>
  dplyr::summarise(do_union = FALSE) |>
  st_cast("LINESTRING")
tracks_tbl <- points_tbl |>
  dplyr::arrange(track_id, timestamp) |>
  dplyr::summarise(
    wkt = paste0(
      "LINESTRING (",
      paste(paste(lon, lat), collapse = ", "),
      ")"
    ),
    .by = track_id
  )
terra_tracks <- terra::vect(tracks_tbl, geom = "wkt", crs = "EPSG:4326")
aoi_qc <- st_transform(aoi, 32198)
zones_qc <- st_transform(zones, 32198)
points_sf_qc <- st_transform(points_sf, 32198)
tracks_sf_qc <- st_transform(tracks_sf, 32198)
terra_aoi_qc <- terra::project(terra_aoi, "EPSG:32198")
terra_zones_qc <- terra::project(terra_zones, "EPSG:32198")
terra_points_qc <- terra::project(terra_points, "EPSG:32198")
terra_tracks_qc <- terra::project(terra_tracks, "EPSG:32198")
focus_area <- zones |>
  dplyr::filter(zone_id %in% c("NE", "SE")) |>
  dplyr::summarise()
points_focus_sf <- st_filter(points_sf, focus_area)
zones_focus_sf <- st_crop(zones, st_bbox(focus_area))
terra_focus_area <- terra_zones[terra_zones$zone_id %in% c("NE", "SE")]
terra_points_focus <- crop(terra_points, terra_focus_area)
terra_zones_focus <- crop(terra_zones, ext(terra_focus_area))
zone_area_sf <- data.frame(
  zone_id = zones_qc$zone_id,
  area_km2 = round(as.numeric(units::set_units(st_area(zones_qc), "km^2")), 1)
)
track_length_sf <- data.frame(
  track_id = tracks_sf_qc$track_id,
  length_km = round(as.numeric(units::set_units(st_length(tracks_sf_qc), "km")), 1)
)
point_boundary_sf <- data.frame(
  obs_id = points_sf_qc$obs_id,
  boundary_km = round(as.numeric(units::set_units(st_distance(points_sf_qc, st_boundary(aoi_qc)), "km")), 1)
)
zone_area_terra <- data.frame(
  zone_id = terra_zones_qc$zone_id,
  area_km2 = round(expanse(terra_zones_qc, unit = "km"), 1)
)
track_length_terra <- data.frame(
  track_id = terra_tracks_qc$track_id,
  length_km = round(perim(terra_tracks_qc) / 1000, 1)
)
point_boundary_terra <- data.frame(
  obs_id = terra_points_qc$obs_id,
  boundary_km = round(distance(terra_points_qc, as.lines(terra_aoi_qc))[, 1] / 1000, 1)
)
points_joined_sf <- st_join(points_sf, zones[, c("zone_id", "zone_type")])
terra_point_attrs <- terra::extract(terra_zones, terra_points)
terra_points_joined <- cbind(terra_points, terra_point_attrs[, c("zone_id", "zone_type")])
track_segments_sf <- suppressWarnings(st_intersection(tracks_sf_qc, zones_qc))
terra_track_segments <- intersect(terra_tracks_qc, terra_zones_qc)
point_buffers_sf <- st_buffer(points_sf_qc, dist = 10000)
terra_point_buffers <- buffer(terra_points_qc, width = 10000)
```



# Welcome 

<!-- Go around the table + slide insileco? -->
<!-- What do you expect  -->

# Objectives

## What To Expect

:::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## What This Workshop Is

- a practical introduction to spatial workflows in R
- a workshop about using R operationally as a GIS
- focused on common manipulations you will reuse in real projects
:::
:::
::: {.column width="50%"}
::: {.callout-warning}
## What This Workshop Is Not

- not a full course in geodesy, cartography, or remote sensing
- not a survey of every spatial package or method
- not a theory-first geocomputation course
:::
:::
::::


## Resources

- [Visualisation et analyse de données spatiales sous R (FR)](https://sci1031.github.io/sci1031/)
- [Geocomputation with R](https://r.geocompx.org/)
- [Spatial Data Science](https://r-spatial.org/book/)

## Workshop Objective

By the end of the workshop, participants should be able to use R as a GIS, work operationally with common spatial data manipulations, and have the right tools in hand to keep learning.

## Our Focus

- import spatial data
- inspect geometry and CRS
- transform and subset data
- measure distance, area, and length
- combine layers through joins and overlays
- extract raster values to vector features
- build summaries, analyses, and maps

## Workshop Structure

:::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## Materials

- short concept-driven slides
- live examples in R
- reusable scripts and resources for later review
:::
:::
::: {.column width="50%"}
::: {.callout-tip}
## Exercises

- targeted hands-on exercises
- designed specifically for workshop participants
- based on the data shared ahead of the event
:::
:::
::::

::: {.callout icon=false}
The workshop alternates between concise instructional material and applied exercises, so participants can move directly from concepts to practice.
:::

# R as a GIS

<!-- Discuss why it makes sense to use R as a GIS -->

## R to build reproducible pipelines

![](img/flow.svg)

::: {.notes}
Could be python etc...
Similar as QGIS for C libraries
:::


## Spatial Data

<table class="spatial-objects-table">
  <thead>
    <tr>
      <th>Object type</th>
      <th>Description</th>
      <th><code>sf</code></th>
      <th><code>stars</code></th>
      <th><code>terra</code></th>
    </tr>
  </thead>
  <tbody>
    <tr class="family-row">
      <td>Vector</td>
      <td class="family-note"><em>Where are the features?</em></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td><img src="img/point.svg" class="obj-icon" alt="Point icon"> Points</td>
      <td>Discrete locations with coordinates and attributes.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr>
      <td><img src="img/line.svg" class="obj-icon" alt="Line icon"> Lines</td>
      <td>Linear features such as roads, rivers, or tracks.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr>
      <td><img src="img/polygon.svg" class="obj-icon" alt="Polygon icon"> Polygons</td>
      <td>Areas such as lakes, study zones, or administrative boundaries.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr class="family-row">
      <td>Raster</td>
      <td class="family-note"><em>What is the value across space?</em></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td><img src="img/grid.svg" class="obj-icon" alt="Raster grid icon"> Rasters</td>
      <td>Gridded surfaces with values stored in cells.</td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
  </tbody>
</table>

## Main Spatial R Packages

***New ecosystem***

:::: {.columns}
::: {.column width="33%"}

::: {.callout-tip icon=false}
### `sf` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/sf/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="sf documentation">&#127760;</a></span>

- *simple features* [standard](https://www.iso.org/standard/40114.html)
- vector data
- tidy friendly
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `stars` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/stars/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="stars documentation">&#127760;</a></span>

- spatiotemporal arrays
- raster and datacubes
- strong companion to `sf`
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `terra` <span class="pkg-link-wrap"><a href="https://rspatial.github.io/terra/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="terra documentation">&#127760;</a></span>

- vector and raster support
- strong raster tooling
- full GIS-style workflow
:::
:::
::::

***Old ecosystem***

:::: {.columns}
::: {.column width="33%"}

::: {.callout-important icon=false}
### `sp` <span class="pkg-link-wrap"><a href="https://cran.r-project.org/web/packages/sp/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="sp documentation">&#127760;</a></span>

- vector data
:::
:::
::: {.column width="33%"}
::: {.callout-important icon=false}
### `raster` <span class="pkg-link-wrap"><a href="https://cran.r-project.org/web/packages/raster/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="raster documentation">&#127760;</a></span>

- rasters
:::
:::
::: {.column width="33%"}
:::
::::

## Accompanying Packages

***Data manipulation and measurements***

:::: {.columns}
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `tidyverse` <span class="pkg-link-wrap"><a href="https://www.tidyverse.org/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tidyverse website">&#127760;</a></span>

- tabular workflows around spatial objects
- filtering, joining, grouping, summarising
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `tidyterra` <span class="pkg-link-wrap"><a href="https://dieghernan.github.io/tidyterra/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tidyterra documentation">&#127760;</a></span>

- tidy verbs and plotting helpers for `terra`
- a useful extension, not a replacement for `terra`
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `units` <span class="pkg-link-wrap"><a href="https://r-quantities.github.io/units/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="units documentation">&#127760;</a></span>

- explicit lengths, areas, and distances in `sf`
- contrasts `terra`, returns numeric measurements
:::
:::
::::

***Visualization***

:::: {.columns}
::: {.column width="33%"}
::: {.callout-note icon=false}
### `ggplot2` <span class="pkg-link-wrap"><a href="https://ggplot2.tidyverse.org/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="ggplot2 website">&#127760;</a></span>

- communication-quality static graphics
- flexible layered maps
:::
:::
::: {.column width="33%"}
::: {.callout-note icon=false}
### `mapview` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/mapview/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="mapview documentation">&#127760;</a></span>

- fast interactive exploration
- quick visual checks of spatial layers
:::
:::
::: {.column width="33%"}
::: {.callout-note icon=false}
### `tmap` <span class="pkg-link-wrap"><a href="https://r-tmap.github.io/tmap/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tmap website">&#127760;</a></span>

- thematic mapping workflows
- static and interactive map support
:::
:::
::::

::: {.callout-warning}
#### These are examples of particularly relevant accompanying packages. The spatial ecosystem in R is broad and active, and is not limited to the packages shown here.
:::

## A Note on IDEs

*What are IDEs? Integrated Development Environments*

:::: {.columns}
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `RStudio` <span class="pkg-link-wrap"><a href="https://posit.co/products/open-source/rstudio/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="RStudio website">&#127760;</a></span>

- long-standing default environment for R
- integrated console, editor, plots, files, and packages
- a strong starting point for most R users
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `VS Code` <span class="pkg-link-wrap"><a href="https://code.visualstudio.com/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="VS Code website">&#127760;</a></span>

- general-purpose editor with R support through extensions
- suited for mixed workflows across R, Python, Quarto, Git
- a good option for multi-language projects
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `Positron` <span class="pkg-link-wrap"><a href="https://positron.posit.co/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="Positron website">&#127760;</a></span>

- newer data-science IDE from Posit, VS Code-like
- combines scripting, notebooks, and IDE features
- worth watching as the ecosystem evolves
:::
:::
::::

::: {.callout-warning}
### The goal of this workshop is not to teach a specific IDE. Use the environment that lets you work comfortably with scripts, objects, plots, and spatial outputs.
:::

## A Note on Pipes 

::: {.callout-note}
### What are pipes?

Pipes pass the output of one step into the next, making multi-step workflows easier to read and write. We mention them explicitly because they are used throughout the workshop.
:::

<br>

:::: {.columns}

::: {.column width="50%"}

`magrittr` ***pipes:*** `%>%`

```R
CO2 %>%
  dplyr::filter(Type == "Quebec")
```
:::

::: {.column width="50%"}

***native pipes:*** `|>`

```R
CO2 |>
  dplyr::filter(Type == "Quebec")
```

:::

::::

## A Note on Code

***Choose Your Tool(s)***

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| eval: false
sf_object <- sf::function(...)
stars_object <- stars::function(...)
```

##### `terra`

```{r}
#| eval: false
terra_object <- terra::function(...)
```
:::

<br> 

::: {.callout-note}
### Navigating through the code in the presentation
The slides often show the same operation twice: once with `sf/stars`, and once with `terra`.

You do not need to follow both versions at once. Use the tabs to stay in the ecosystem you chose for the workshop.
:::



## A Note on Exercises

***Choose Your Route***

![](img/route_matrix.svg){width=82% fig-align="center"}

::: {.callout-note}
### Do both levels!
More advanced participants can complete both workflows, as there will be advanced workflows available with both points and lines. 
:::


# Vector Data in R

***sf*** & ***terra***

## Read Vector Data

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
aoi <- st_read("data/polygons.gpkg", layer = "study_area")
zones <- st_read("data/polygons.gpkg", layer = "zones")
aoi
zones
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(aoi)
head(zones)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
#| 
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2.5,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones),
  add = TRUE,
  col = c("#9ecae1aa", "#fdae6baa", "#a1d99baa", "#fdd0a2aa"),
  border = "#264653",
  lwd = 1.2
)
zone_centers <- st_coordinates(st_centroid(zones))
text(
  zone_centers[, 1],
  zone_centers[, 2],
  labels = zones$zone_id,
  cex = 0.9,
  font = 2,
  col = "#17323a"
)
```

:::
:::
<!-- end columns -->


##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
aoi <- vect("data/polygons.gpkg", layer = "study_area")
zones <- vect("data/polygons.gpkg", layer = "zones")
aoi
zones
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_aoi)
head(terra_zones)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2.5,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones,
  add = TRUE,
  col = c("#9ecae1aa", "#fdae6baa", "#a1d99baa", "#fdd0a2aa"),
  border = "#264653",
  lwd = 1.2
)
zone_centers <- terra::crds(terra::centroids(terra_zones))
text(
  zone_centers[, 1],
  zone_centers[, 2],
  labels = terra_zones$zone_id,
  cex = 0.9,
  font = 2,
  col = "#17323a"
)
```

:::
:::
<!-- end columns -->
:::

## Create Points From A Table

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
stations <- read.csv("data/points.csv") |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326, remove = FALSE)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(points_sf)
```
:::
::: {.column width="35%"}
<br>

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones),
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  st_geometry(points_sf),
  add = TRUE,
  pch = 21,
  bg = "#c44e5288",
  col = "#7f1d1d",
  cex = 1.1
)
```

:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
stations <- read.csv("data/points.csv") |>
  vect(geom = c("lon", "lat"), crs = "EPSG:4326")
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_points)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones,
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  terra_points,
  add = TRUE,
  pch = 21,
  bg = "#c44e5288",
  col = "#7f1d1d",
  cex = 1.1
)
```

:::
:::
<!-- end columns -->
:::


## Inspect Vector Data

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
st_geometry_type(zones)
st_crs(zones)$input
st_bbox(zones)
```
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
st_geometry_type(zones)
st_crs(zones)$input
st_bbox(zones)
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
geomtype(terra_zones)
crs(terra_zones, proj = TRUE)
ext(terra_zones)
```
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
geomtype(terra_zones)
crs(terra_zones, proj = TRUE)
ext(terra_zones)
```
:::

<br> 

::: {.callout-note}
### Spatial projections
CRS information tells you how coordinates are referenced on the Earth. When you need to inspect or identify a projection, [epsg.io](https://epsg.io/) is a useful lookup tool.
:::

## Transform CRS

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
zones_qc <- st_transform(zones, 32198)
points_qc <- st_transform(points_sf, 32198)
st_crs(points_qc)$input
st_bbox(points_qc)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
st_crs(points_sf_qc)$input
st_bbox(points_sf_qc)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi_qc),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones_qc),
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  st_geometry(points_sf_qc),
  add = TRUE,
  pch = 21,
  bg = "#c44e5288",
  col = "#7f1d1d",
  cex = 1.1
)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
zones_qc <- project(terra_zones, "EPSG:32198")
points_qc <- project(terra_points, "EPSG:32198")
crs(points_qc, proj = TRUE)
ext(points_qc)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
crs(terra_points_qc, proj = TRUE)
ext(terra_points_qc)
```
:::
::: {.column width="35%"}

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi_qc,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones_qc,
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  terra_points_qc,
  add = TRUE,
  pch = 21,
  bg = "#c44e5288",
  col = "#7f1d1d",
  cex = 1.1
)
```
:::
:::
<!-- end columns -->
:::

::: {.callout-warning}
## Measurements 
Use a suitable projected CRS when you need interpretable distances, lengths, or areas.
:::

## Export Vector Outputs

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
st_write(points_sf_qc, "outputs/points_projected.gpkg")
```

```{r}
#| message: false
#| warning: false
#| echo: false
sf_export <- tempfile(fileext = ".gpkg")
st_write(points_sf_qc, sf_export, quiet = TRUE)
data.frame(
  object = "points_sf_qc",
  file = basename(sf_export),
  exists = file.exists(sf_export),
  features = nrow(points_sf_qc)
)
```

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
writeVector(terra_points_qc, "outputs/points_projected.gpkg")
```

```{r}
#| message: false
#| warning: false
#| echo: false
terra_export <- tempfile(fileext = ".gpkg")
writeVector(terra_points_qc, terra_export, overwrite = TRUE)
data.frame(
  object = "terra_points_qc",
  file = basename(terra_export),
  exists = file.exists(terra_export),
  features = nrow(terra_points_qc)
)
```
:::

## Exercise 1 {.exercise-break .center}

### Vector Foundations

Import, inspect, transform, export

<br>

*Bonus: use R as a converter*

## Build Lines From Ordered Points

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
tracks <- points_sf |>
  dplyr::arrange(track_id, timestamp) |>
  dplyr::group_by(track_id) |>
  dplyr::summarise(do_union = FALSE) |>
  st_cast("LINESTRING")
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(tracks_sf)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones),
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  st_geometry(tracks_sf),
  add = TRUE,
  col = "#7f1d1d",
  lwd = 2.3
)
plot(
  st_geometry(points_sf),
  add = TRUE,
  pch = 16,
  col = "#7f1d1daa",
  cex = 0.65
)
```

:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
tracks_tbl <- points_tbl |>
  dplyr::arrange(track_id, timestamp) |>
  dplyr::summarise(
    wkt = paste0(
      "LINESTRING (",
      paste(paste(lon, lat), collapse = ", "),
      ")"
    ),
    .by = track_id
  ) |>
  terra::vect(geom = "wkt", crs = "EPSG:4326")
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_tracks)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones,
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  terra_tracks,
  add = TRUE,
  col = "#7f1d1d",
  lwd = 2.3
)
plot(
  terra_points,
  add = TRUE,
  pch = 16,
  col = "#7f1d1daa",
  cex = 0.65
)
```

:::
:::
<!-- end columns -->
:::

## Quick Static Mapping

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| eval: false
plot(st_geometry(aoi), col = "#f7fbfc", border = "#355c67")
plot(st_geometry(zones), add = TRUE, border = "#264653")
plot(st_geometry(stations), pch = 16, col = "#7f1d1d", add = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center
#| out-width: 60%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  st_geometry(zones),
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  st_geometry(points_sf),
  add = TRUE,
  pch = 16,
  col = "#7f1d1d",
  cex = 0.9
)
```

##### `terra`

```{r}
#| eval: false
plot(aoi, col = "#f7fbfc", border = "#355c67")
plot(zones, add = TRUE, border = "#264653")
plot(stations, pch = 16, col = "#7f1d1d", add = TRUE)
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| fig-align: center
#| out-width: 60%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(
  terra_zones,
  add = TRUE,
  col = c("#9ecae144", "#fdae6b44", "#a1d99b44", "#fdd0a244"),
  border = "#264653",
  lwd = 1
)
plot(
  terra_points,
  add = TRUE,
  pch = 16,
  col = "#7f1d1d",
  cex = 0.9
)
```
:::

## Quick Interactive Mapping

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| eval: false
mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(stations, zcol = "category", layer.name = "Observations")
```

```{r}
#| message: false
#| warning: false
#| echo: false
mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(points_sf, zcol = "category", layer.name = "Observations")
```

##### `terra`

```{r}
#| eval: false
mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(stations, zcol = "category", layer.name = "Observations")
```

```{r}
#| message: false
#| warning: false
#| echo: false
mapview(terra_zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(terra_points, zcol = "category", layer.name = "Observations")
```
:::

## Export Interactive Maps

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| eval: false
m <- mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(stations, zcol = "category", layer.name = "Observations")

mapview::mapshot(m, file = "outputs/interactive_map.html")
```

```{r}
#| message: false
#| warning: false
#| echo: false
sf_map_file <- tempfile(fileext = ".html")
data.frame(
  object = "mapview sf map",
  file = basename(sf_map_file),
  format = "html"
)
```

##### `terra`

```{r}
#| eval: false
m <- mapview(zones, zcol = "zone_id", layer.name = "Zones") +
  mapview(stations, zcol = "category", layer.name = "Observations")

mapview::mapshot(m, file = "outputs/interactive_map.html")
```

```{r}
#| message: false
#| warning: false
#| echo: false
terra_map_file <- tempfile(fileext = ".html")
data.frame(
  object = "mapview terra map",
  file = basename(terra_map_file),
  format = "html"
)
```
:::

<br>

::: {.callout-note}
### Share your maps!
Interactive `mapview` outputs can be saved as standalone HTML files and shared like any other web document.
:::

## Exercise 2 {.exercise-break .center}

### Vector Foundations on Your Data

Import, inspect, transform, export, visualize


## Subset To A Study Area

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
focus_area <- zones |>
  dplyr::filter(zone_id %in% c("NE", "SE")) |>
  dplyr::summarise()

points_focus <- st_filter(points_sf, focus_area)
zones_focus <- st_crop(zones, st_bbox(focus_area))
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(points_focus_sf)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  st_geometry(aoi),
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(st_geometry(focus_area), add = TRUE, col = "#37abc833", border = "#37abc8", lwd = 2)
plot(st_geometry(points_sf), add = TRUE, pch = 16, col = "#9aa5ab88", cex = 0.7)
plot(st_geometry(points_focus_sf), add = TRUE, pch = 21, bg = "#c44e5288", col = "#7f1d1d", cex = 1.1)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
focus_area <- terra_zones[terra_zones$zone_id %in% c("NE", "SE")]
points_focus <- crop(terra_points, focus_area)
zones_focus <- crop(terra_zones, ext(focus_area))
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_points_focus)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(
  terra_aoi,
  col = "#f7fbfc",
  border = "#355c67",
  lwd = 2,
  axes = FALSE,
  main = ""
)
plot(terra_focus_area, add = TRUE, col = "#37abc833", border = "#37abc8", lwd = 2)
plot(terra_points, add = TRUE, pch = 16, col = "#9aa5ab88", cex = 0.7)
plot(terra_points_focus, add = TRUE, pch = 21, bg = "#c44e5288", col = "#7f1d1d", cex = 1.1)
```
:::
:::
<!-- end columns -->
:::

## Measure Vector Features

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
zone_area <- units::set_units(st_area(zones_qc), "km^2")
track_length <- units::set_units(st_length(tracks_sf_qc), "km")
boundary_distance <- units::set_units(st_distance(points_sf_qc, st_boundary(aoi_qc)), "km")
zone_area
track_length
head(boundary_distance)
```

::: {.columns}
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
zone_area_sf
track_length_sf
```
:::
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
head(point_boundary_sf)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
zone_area <- expanse(terra_zones_qc, unit = "km")
track_length <- perim(terra_tracks_qc) / 1000
boundary_distance <- distance(terra_points_qc, as.lines(terra_aoi_qc))[, 1] / 1000
zone_area
track_length
head(boundary_distance)
```

::: {.columns}
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
zone_area_terra
track_length_terra
```
:::
::: {.column width="50%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| results: markup
head(point_boundary_terra)
```
:::
:::
<!-- end columns -->

:::

<br>

::: {.callout-note icon=false}
`sf` often returns unit-aware measurements. `terra` often returns numeric values whose meaning depends on the CRS and function used.
:::

## Join Attributes To Features

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
points_joined <- st_join(points_sf, zones[, c("zone_id", "zone_type")])
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(sf::st_drop_geometry(points_joined_sf))
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
zone_cols <- c(NW = "#2c7fb8", NE = "#f03b20", SW = "#31a354", SE = "#fdae61")
par(mar = c(0, 0, 0, 0))
plot(st_geometry(aoi), col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(st_geometry(zones), add = TRUE, col = paste0(zone_cols[zones$zone_id], "33"), border = "#264653", lwd = 1)
plot(st_geometry(points_joined_sf), add = TRUE, pch = 21, bg = zone_cols[points_joined_sf$zone_id], col = "#17323a", cex = 1.1)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
point_values <- extract(terra_zones, terra_points)
points_joined <- cbind(terra_points, point_values[, c("zone_id", "zone_type")])
```

::: {.columns}
::: {.column width="68%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_points_joined)
```
:::
::: {.column width="32%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
zone_cols <- c(NW = "#2c7fb8", NE = "#f03b20", SW = "#31a354", SE = "#fdae61")
par(mar = c(0, 0, 0, 0))
plot(terra_aoi, col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(terra_zones, add = TRUE, col = paste0(zone_cols[terra_zones$zone_id], "33"), border = "#264653", lwd = 1)
plot(terra_points_joined, add = TRUE, pch = 21, bg = zone_cols[terra_points_joined$zone_id], col = "#17323a", cex = 1.1)
```
:::
:::
<!-- end columns -->
:::

## Intersect Geometries

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
track_segments <- st_intersection(tracks_sf_qc, zones_qc)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(sf::st_drop_geometry(track_segments_sf))
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
zone_cols <- c(NW = "#2c7fb8", NE = "#f03b20", SW = "#31a354", SE = "#fdae61")
par(mar = c(0, 0, 0, 0))
plot(st_geometry(aoi_qc), col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(st_geometry(zones_qc), add = TRUE, col = paste0(zone_cols[zones_qc$zone_id], "22"), border = "#264653", lwd = 1)
plot(st_geometry(track_segments_sf), add = TRUE, col = zone_cols[track_segments_sf$zone_id], lwd = 2.5)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
track_segments <- intersect(terra_tracks_qc, terra_zones_qc)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
head(terra_track_segments)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
zone_cols <- c(NW = "#2c7fb8", NE = "#f03b20", SW = "#31a354", SE = "#fdae61")
par(mar = c(0, 0, 0, 0))
plot(terra_aoi_qc, col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(terra_zones_qc, add = TRUE, col = paste0(zone_cols[terra_zones_qc$zone_id], "22"), border = "#264653", lwd = 1)
plot(terra_track_segments, add = TRUE, col = zone_cols[terra_track_segments$zone_id], lwd = 2.5)
```
:::
:::
<!-- end columns -->
:::

::: {.callout-note icon=false}
Use a join when you want to attach attributes. Use an intersection when you want overlap to create new geometries.
:::

## Buffers

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| message: false
#| warning: false
#| eval: false
point_buffers <- st_buffer(points_sf_qc, dist = 10000)
st_geometry_type(point_buffers)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
st_geometry_type(point_buffers_sf)[1]
head(sf::st_drop_geometry(point_buffers_sf))
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(st_geometry(aoi_qc), col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(st_geometry(zones_qc), add = TRUE, col = "#9ecae122", border = "#264653", lwd = 1)
plot(st_geometry(point_buffers_sf), add = TRUE, col = "#37abc833", border = "#37abc8", lwd = 1)
plot(st_geometry(points_sf_qc), add = TRUE, pch = 16, col = "#7f1d1d", cex = 0.7)
```
:::
:::
<!-- end columns -->

##### `terra`

```{r}
#| message: false
#| warning: false
#| eval: false
point_buffers <- buffer(terra_points_qc, width = 10000)
geomtype(point_buffers)
```

::: {.columns}
::: {.column width="65%"}
```{r}
#| message: false
#| warning: false
#| echo: false
geomtype(terra_point_buffers)
head(terra_point_buffers)
```
:::
::: {.column width="35%"}
```{r}
#| message: false
#| warning: false
#| echo: false
#| fig-width: 7
#| fig-height: 5.5
#| out-width: 100%
par(mar = c(0, 0, 0, 0))
plot(terra_aoi_qc, col = "#f7fbfc", border = "#355c67", lwd = 2, axes = FALSE, main = "")
plot(terra_zones_qc, add = TRUE, col = "#9ecae122", border = "#264653", lwd = 1)
plot(terra_point_buffers, add = TRUE, col = "#37abc833", border = "#37abc8", lwd = 1)
plot(terra_points_qc, add = TRUE, pch = 16, col = "#7f1d1d", cex = 0.7)
```
:::
:::
<!-- end columns -->
:::

::: {.callout-note icon=false}
Buffers create new geometries at a fixed distance around features, so they are usually done in a projected CRS with interpretable distance units.
:::

## Exercise 3 {.exercise-break .center}

### Vector Operations

Subset, measure, join, intersect, buffer


# Raster Data

## Read Raster Data

:::: {.columns}
::: {.column width="50%"}
`stars`

```{r}
#| eval: false
elevation <- read_stars("elevation.tif")
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
elevation <- rast("elevation.tif")
```
:::
::::

## Inspect Raster Data

- how many layers?
- what is the resolution?
- what is the extent?
- what CRS is it in?
- what do the values represent?

## Basic Raster Plotting

- use `plot()` for a quick check
- inspect the color scale and coverage
- compare raster extent to vector layers

## Crop, Mask, Project

:::: {.columns}
::: {.column width="33%"}
`Crop`

- reduce the spatial extent
- keep a window of interest
:::
::: {.column width="33%"}
`Mask`

- keep cells only inside a geometry
- remove cells outside a polygon
:::
::: {.column width="33%"}
`Project`

- change CRS
- usually implies resampling
:::
::::

## Raster Transformation API Lens

:::: {.columns}
::: {.column width="50%"}
`sf/stars`

```{r}
#| eval: false
elevation_lcc <- st_warp(elevation, crs = st_crs(city_boundary))
elevation_city <- st_crop(elevation_lcc, st_bbox(city_boundary))
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
elevation_lcc <- project(elevation, "EPSG:32198")
elevation_city <- crop(elevation_lcc, ext(city_boundary))
```
:::
::::

## Raster Questions To Ask

- do I need to reproject this raster?
- should I crop first or project first?
- does the current resolution fit my question?
- am I using this raster for display or analysis?

# Integrated Workflows

## Why Integration Matters

- vector features define locations, paths, and areas
- rasters provide continuous environmental context
- many analyses require combining both

## Extract Raster Values To Vectors

- points: value at each location
- polygons: mean, min, max, or sum within each area
- lines: summaries along roads, rivers, or tracks

## Extraction API Lens

:::: {.columns}
::: {.column width="50%"}
`sf/stars`

```{r}
#| eval: false
stations$temp <- st_extract(temperature, stations)[[1]][, 1]
zones$mean_elev <- st_extract(elevation, zones, FUN = mean)[[1]]
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
station_vals <- extract(temperature, stations)
zone_vals <- extract(elevation, zones, fun = mean)
```
:::
::::

## Build Analysis Tables

- drop geometry when the spatial step is complete
- keep only variables needed for analysis
- summarise by feature, zone, or category

## Table-Building Verbs

- `select()`
- `mutate()`
- `group_by()`
- `summarise()`
- `count()`
- `left_join()`

## From Spatial Data To Models

- count events by area
- compare mean raster values across zones
- model a response as a function of spatial predictors
- use spatial outputs as inputs to standard statistical tools

## A Simple Analysis Pattern

1. create or import features
2. align CRS
3. subset to a study area
4. measure or overlay vectors
5. extract raster values
6. summarise into a clean table
7. analyse or visualize

## KDE As An Exploratory Tool

- useful for point patterns
- produces a smooth intensity surface
- helps visualize concentration or hotspots
- often used for exploration rather than formal inference

## Static And Interactive Mapping

:::: {.columns}
::: {.column width="50%"}
`Quick exploration`

- `plot()`
- `mapview()`
:::
::: {.column width="50%"}
`Communication`

- `ggplot2`
- careful legends, colors, labels, and layers
:::
::::

## Export Outputs

:::: {.columns}
::: {.column width="50%"}
`Vector`

- `st_write()`
- `writeVector()`
:::
::: {.column width="50%"}
`Tables`

- `write.csv()`
- `readr::write_csv()`
:::
::::

## A Practical Workflow Mindset

- inspect early
- transform deliberately
- keep data in compatible CRS
- separate exploratory maps from final figures
- move to plain tables when the spatial step is done

## Takeaways

- vector and raster are different data models
- `sf`, `stars`, and `terra` cover most common spatial workflows in R
- CRS decisions affect measurements and overlays
- joins, intersections, crops, masks, and extraction are core operations
- spatial analysis in R is often a bridge between GIS tasks and standard statistical workflows

## Closing

- start simple
- verify geometry and CRS often
- build reusable workflows
- treat R as both an analysis environment and a GIS
