---
title: "R in Space"
author: "*inSileco*"
date: "2026-03-25"
from: markdown+emoji
---

```{=html}
<script>
document.addEventListener('DOMContentLoaded', function() {
  var logo = document.querySelector('.slide-logo');
  if (logo) {
    var link = document.createElement('a');
    link.href = 'https://www.insileco.io';
    link.target = '_blank';
    logo.parentNode.insertBefore(link, logo);
    link.appendChild(logo);
  }
});
</script>
```

# Welcome 

<!-- Go around the table + slide insileco? -->
<!-- What do you expect  -->

# Objectives

## What To Expect

:::: {.columns}
::: {.column width="50%"}
::: {.callout-note}
## What This Workshop Is

- a practical introduction to spatial workflows in R
- a workshop about using R operationally as a GIS
- focused on common manipulations you will reuse in real projects
:::
:::
::: {.column width="50%"}
::: {.callout-warning}
## What This Workshop Is Not

- not a full course in geodesy, cartography, or remote sensing
- not a survey of every spatial package or method
- not a theory-first geocomputation course
:::
:::
::::


## Resources

- [Visualisation et analyse de données spatiales sous R (FR)](https://sci1031.github.io/sci1031/)
- [Geocomputation with R](https://r.geocompx.org/)
- [Spatial Data Science](https://r-spatial.org/book/)

## Workshop Objective

By the end of the workshop, participants should be able to use R as a GIS, work operationally with common spatial data manipulations, and have the right tools in hand to keep learning.

## Our Focus

- import spatial data
- inspect geometry and CRS
- transform and subset data
- measure distance, area, and length
- combine layers through joins and overlays
- extract raster values to vector features
- build summaries, analyses, and maps

# R as a GIS

<!-- Discuss why it makes sense to use R -->

## R to build reproducible pipelines

![](img/flow.svg)

::: {.notes}
Could be python etc...
Similar as QGIS for C libraries
:::


## Spatial Data

<table class="spatial-objects-table">
  <thead>
    <tr>
      <th>Object type</th>
      <th>Description</th>
      <th><code>sf</code></th>
      <th><code>stars</code></th>
      <th><code>terra</code></th>
    </tr>
  </thead>
  <tbody>
    <tr class="family-row">
      <td>Vector</td>
      <td class="family-note"><em>Where are the features?</em></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td><img src="img/point.svg" class="obj-icon" alt="Point icon"> Points</td>
      <td>Discrete locations with coordinates and attributes.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr>
      <td><img src="img/line.svg" class="obj-icon" alt="Line icon"> Lines</td>
      <td>Linear features such as roads, rivers, or tracks.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr>
      <td><img src="img/polygon.svg" class="obj-icon" alt="Polygon icon"> Polygons</td>
      <td>Areas such as lakes, study zones, or administrative boundaries.</td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
    <tr class="family-row">
      <td>Raster</td>
      <td class="family-note"><em>What is the value across space?</em></td>
      <td></td>
      <td></td>
      <td></td>
    </tr>
    <tr>
      <td><img src="img/grid.svg" class="obj-icon" alt="Raster grid icon"> Rasters</td>
      <td>Gridded surfaces with values stored in cells.</td>
      <td></td>
      <td><span class="pkg-check">&#10003;</span></td>
      <td><span class="pkg-check">&#10003;</span></td>
    </tr>
  </tbody>
</table>

## Main Spatial R Packages

***New ecosystem***

:::: {.columns}
::: {.column width="33%"}

::: {.callout-tip icon=false}
### `sf` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/sf/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="sf documentation">&#127760;</a></span>

- *simple features* [standard](https://www.iso.org/standard/40114.html)
- vector data
- tidy friendly
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `stars` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/stars/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="stars documentation">&#127760;</a></span>

- spatiotemporal arrays
- raster and datacubes
- strong companion to `sf`
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `terra` <span class="pkg-link-wrap"><a href="https://rspatial.github.io/terra/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="terra documentation">&#127760;</a></span>

- vector and raster support
- strong raster tooling
- full GIS-style workflow
:::
:::
::::

***Old ecosystem***

:::: {.columns}
::: {.column width="33%"}

::: {.callout-important icon=false}
### `sp` <span class="pkg-link-wrap"><a href="https://cran.r-project.org/web/packages/sp/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="sp documentation">&#127760;</a></span>

- vector data
:::
:::
::: {.column width="33%"}
::: {.callout-important icon=false}
### `raster` <span class="pkg-link-wrap"><a href="https://cran.r-project.org/web/packages/raster/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="raster documentation">&#127760;</a></span>

- rasters
:::
:::
::: {.column width="33%"}
:::
::::

## Accompanying Packages

***Data manipulation and measurements***

:::: {.columns}
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `tidyverse` <span class="pkg-link-wrap"><a href="https://www.tidyverse.org/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tidyverse website">&#127760;</a></span>

- tabular workflows around spatial objects
- filtering, joining, grouping, summarising
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `tidyterra` <span class="pkg-link-wrap"><a href="https://dieghernan.github.io/tidyterra/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tidyterra documentation">&#127760;</a></span>

- tidy verbs and plotting helpers for `terra`
- a useful extension, not a replacement for `terra`
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `units` <span class="pkg-link-wrap"><a href="https://r-quantities.github.io/units/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="units documentation">&#127760;</a></span>

- explicit lengths, areas, and distances in `sf`
- contrasts `terra`, returns numeric measurements
:::
:::
::::

***Visualization***

:::: {.columns}
::: {.column width="33%"}
::: {.callout-note icon=false}
### `ggplot2` <span class="pkg-link-wrap"><a href="https://ggplot2.tidyverse.org/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="ggplot2 website">&#127760;</a></span>

- communication-quality static graphics
- flexible layered maps
:::
:::
::: {.column width="33%"}
::: {.callout-note icon=false}
### `mapview` <span class="pkg-link-wrap"><a href="https://r-spatial.github.io/mapview/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="mapview documentation">&#127760;</a></span>

- fast interactive exploration
- quick visual checks of spatial layers
:::
:::
::: {.column width="33%"}
::: {.callout-note icon=false}
### `tmap` <span class="pkg-link-wrap"><a href="https://r-tmap.github.io/tmap/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="tmap website">&#127760;</a></span>

- thematic mapping workflows
- static and interactive map support
:::
:::
::::

::: {.callout-warning}
#### These are examples of particularly relevant accompanying packages. The spatial ecosystem in R is broad and active, and is not limited to the packages shown here.
:::

## A Note on IDEs

*What are IDEs? Integrated Development Environments*

:::: {.columns}
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `RStudio` <span class="pkg-link-wrap"><a href="https://posit.co/products/open-source/rstudio/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="RStudio website">&#127760;</a></span>

- long-standing default environment for R
- integrated console, editor, plots, files, and packages
- a strong starting point for most R users
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `VS Code` <span class="pkg-link-wrap"><a href="https://code.visualstudio.com/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="VS Code website">&#127760;</a></span>

- general-purpose editor with R support through extensions
- suited for mixed workflows across R, Python, Quarto, Git
- a good option for multi-language projects
:::
:::
::: {.column width="33%"}
::: {.callout-tip icon=false}
### `Positron` <span class="pkg-link-wrap"><a href="https://positron.posit.co/" target="_blank" rel="noopener noreferrer" class="pkg-link" aria-label="Positron website">&#127760;</a></span>

- newer data-science IDE from Posit, VS Code-like
- combines scripting, notebooks, and IDE features
- worth watching as the ecosystem evolves
:::
:::
::::

::: {.callout-warning}
### The goal of this workshop is not to teach a specific IDE. Use the environment that lets you work comfortably with scripts, objects, plots, and spatial outputs.
:::

## A note on pipes 

::: {.callout-note}
### What are pipes?

Pipes pass the output of one step into the next, making multi-step workflows easier to read and write. We mention them explicitly because they are used throughout the workshop.
:::

<br>

:::: {.columns}

::: {.column width="50%"}

`magrittr` ***pipes:*** `%>%`

```R
CO2 %>%
  dplyr::filter(Type == "Quebec")
```
:::

::: {.column width="50%"}

***native pipes:*** `|>`

```R
CO2 |>
  dplyr::filter(Type == "Quebec")
```

:::

::::




# Vector Data in R

***sf*** & ***terra***

## Create Points From A Table

::: {.panel-tabset .code-tabs}
##### `sf`

```{r}
#| eval: false
stations <- read.csv("stations.csv") |>
  st_as_sf(coords = c("lon", "lat"),
           crs = 4326,
           remove = FALSE)
```

##### `terra`

```{r}
#| eval: false
stations_tbl <- read.csv("stations.csv")
stations <- vect(stations_tbl,
                 geom = c("lon", "lat"),
                 crs = "EPSG:4326")
```
:::

## Build Lines From Ordered Points

- start from point observations with an ID and order variable
- sort points
- group them into trajectories or paths
- convert grouped coordinates into line geometry

## Read Vector Data

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
roads <- st_read("roads.gpkg")
municipalities <- st_read("municipalities.gpkg")
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
roads <- vect("roads.gpkg")
municipalities <- vect("municipalities.gpkg")
```
:::
::::

## Inspect Vector Data

- geometry type
- number of features
- attributes
- CRS
- spatial extent

## CRS: The Practical Questions

- what CRS is this layer in?
- are my layers using the same CRS?
- do I need a projected CRS for measurement?
- which CRS should I use for my study area?

## Inspect And Transform CRS

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
st_crs(stations)
stations_lcc <- st_transform(stations, 32198)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
crs(stations)
stations_lcc <- project(stations, "EPSG:32198")
```
:::
::::

::: {.callout-caution}
Spatial measurements depend on CRS. Distances and areas are easiest to interpret in an appropriate projected CRS.
:::

## Subset To A Study Area

- crop or filter features to a region of interest
- reduce clutter and computation
- align the data to the analysis question

## Vector Subsetting

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
stations_city <- st_filter(stations, city_boundary)
roads_city <- st_crop(roads, st_bbox(city_boundary))
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
stations_city <- crop(stations, city_boundary)
roads_city <- crop(roads, ext(city_boundary))
```
:::
::::

## Measure Vector Features

- area of polygons
- length of lines
- distance between features

## Measurements In Practice

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
lakes$area <- st_area(lakes)
roads$length_m <- st_length(roads)
stations$dist_to_river <- st_distance(stations, rivers)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
lakes$area_km2 <- expanse(lakes, unit = "km")
roads$length_m <- perim(roads)
stations$dist_to_river_m <- distance(stations, rivers)[, 1]
```
:::
::::

::: {.callout-note}
`sf` often uses `units` objects. `terra` usually returns numeric values whose meaning depends on the CRS and function used.
:::

## Join And Overlay

- attach polygon attributes to points
- identify which features fall in which zones
- summarise observations by area

## Intersection

- create new geometries from overlap
- split features where geometries meet
- useful for road segments by municipality, tracks by habitat, or rivers by watershed

## Join And Intersection API Lens

:::: {.columns}
::: {.column width="50%"}
`sf`

```{r}
#| eval: false
stations <- st_join(stations, municipalities)
road_segments <- st_intersection(roads, municipalities)
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
station_values <- extract(municipalities, stations)
road_segments <- intersect(roads, municipalities)
```
:::
::::

## Quick Vector Mapping

- `plot()` for fast checks
- `mapview()` for interactive browsing
- `ggplot2` for more controlled graphics

# Raster Data

## Read Raster Data

:::: {.columns}
::: {.column width="50%"}
`stars`

```{r}
#| eval: false
elevation <- read_stars("elevation.tif")
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
elevation <- rast("elevation.tif")
```
:::
::::

## Inspect Raster Data

- how many layers?
- what is the resolution?
- what is the extent?
- what CRS is it in?
- what do the values represent?

## Basic Raster Plotting

- use `plot()` for a quick check
- inspect the color scale and coverage
- compare raster extent to vector layers

## Crop, Mask, Project

:::: {.columns}
::: {.column width="33%"}
`Crop`

- reduce the spatial extent
- keep a window of interest
:::
::: {.column width="33%"}
`Mask`

- keep cells only inside a geometry
- remove cells outside a polygon
:::
::: {.column width="33%"}
`Project`

- change CRS
- usually implies resampling
:::
::::

## Raster Transformation API Lens

:::: {.columns}
::: {.column width="50%"}
`sf/stars`

```{r}
#| eval: false
elevation_lcc <- st_warp(elevation, crs = st_crs(city_boundary))
elevation_city <- st_crop(elevation_lcc, st_bbox(city_boundary))
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
elevation_lcc <- project(elevation, "EPSG:32198")
elevation_city <- crop(elevation_lcc, ext(city_boundary))
```
:::
::::

## Raster Questions To Ask

- do I need to reproject this raster?
- should I crop first or project first?
- does the current resolution fit my question?
- am I using this raster for display or analysis?

# Integrated Workflows

## Why Integration Matters

- vector features define locations, paths, and areas
- rasters provide continuous environmental context
- many analyses require combining both

## Extract Raster Values To Vectors

- points: value at each location
- polygons: mean, min, max, or sum within each area
- lines: summaries along roads, rivers, or tracks

## Extraction API Lens

:::: {.columns}
::: {.column width="50%"}
`sf/stars`

```{r}
#| eval: false
stations$temp <- st_extract(temperature, stations)[[1]][, 1]
zones$mean_elev <- st_extract(elevation, zones, FUN = mean)[[1]]
```
:::
::: {.column width="50%"}
`terra`

```{r}
#| eval: false
station_vals <- extract(temperature, stations)
zone_vals <- extract(elevation, zones, fun = mean)
```
:::
::::

## Build Analysis Tables

- drop geometry when the spatial step is complete
- keep only variables needed for analysis
- summarise by feature, zone, or category

## Table-Building Verbs

- `select()`
- `mutate()`
- `group_by()`
- `summarise()`
- `count()`
- `left_join()`

## From Spatial Data To Models

- count events by area
- compare mean raster values across zones
- model a response as a function of spatial predictors
- use spatial outputs as inputs to standard statistical tools

## A Simple Analysis Pattern

1. create or import features
2. align CRS
3. subset to a study area
4. measure or overlay vectors
5. extract raster values
6. summarise into a clean table
7. analyse or visualize

## KDE As An Exploratory Tool

- useful for point patterns
- produces a smooth intensity surface
- helps visualize concentration or hotspots
- often used for exploration rather than formal inference

## Static And Interactive Mapping

:::: {.columns}
::: {.column width="50%"}
`Quick exploration`

- `plot()`
- `mapview()`
:::
::: {.column width="50%"}
`Communication`

- `ggplot2`
- careful legends, colors, labels, and layers
:::
::::

## Export Outputs

:::: {.columns}
::: {.column width="50%"}
`Vector`

- `st_write()`
- `writeVector()`
:::
::: {.column width="50%"}
`Tables`

- `write.csv()`
- `readr::write_csv()`
:::
::::

## A Practical Workflow Mindset

- inspect early
- transform deliberately
- keep data in compatible CRS
- separate exploratory maps from final figures
- move to plain tables when the spatial step is done

## Takeaways

- vector and raster are different data models
- `sf`, `stars`, and `terra` cover most common spatial workflows in R
- CRS decisions affect measurements and overlays
- joins, intersections, crops, masks, and extraction are core operations
- spatial analysis in R is often a bridge between GIS tasks and standard statistical workflows

## Closing

- start simple
- verify geometry and CRS often
- build reusable workflows
- treat R as both an analysis environment and a GIS
